---
title: "Spatial RSV Renewal — Stan (FSV rho-est AR(2), Beta(2,2) prior)"
subtitle: "Single-fit: rho estimated jointly, vector[P] σ, AR(2) on lambda with fixed (phi1, phi2) = (2, -1)"
author: "Jose Palacio"
date: today
date-format: "MMMM D, YYYY"
format:
  revealjs:
    theme: simple
    slide-number: true
    progress: true
    code-line-numbers: true
    highlight-style: github
    footer: "Spatial RSV Renewal · Stan · 15 plants · FSV rho-est AR(2) · Beta(2,2)"
    transition: fade
    background-transition: fade
    incremental: false
    toc: false
    smaller: true
    scrollable: true
    fig-align: center
    fig-width: 9
    fig-height: 5
execute:
  echo: true
  eval: true
  warning: false
  message: false
  cache: false
knitr:
  opts_chunk:
    comment: "#>"
editor_options:
  chunk_output_type: console
editor:
  markdown:
    wrap: 72
---

```{=html}
<style>
:root { --font-scale: 0.75; --code-scale: 0.75; }
.reveal .slides         { font-size: calc(40px * var(--font-scale)); }
.reveal .slides h1      { font-size: calc(2.2em * var(--font-scale)); }
.reveal .slides h2      { font-size: calc(1.6em * var(--font-scale)); }
.reveal pre, .reveal pre code { font-size: calc(0.75em * var(--code-scale)); }
.reveal table           { font-size: calc(0.8em * var(--font-scale)); }
</style>
```

```{r}
#| label: root-dir
#| include: false
# proj_root: ruta absoluta a la carpeta donde viven data/, stan/, R/.
# Local Mac: usa absoluta para que funcione desde cualquier cwd.
# NOTS: cambiar a `normalizePath(".")` (cwd del SLURM submit dir).
proj_root <- if (Sys.info()["sysname"] == "Darwin") {
  "/Users/josepalacio/Documents/Claude/Projects/NOTS/rsv_mcmc_stan_geoW"
} else {
  normalizePath(".", mustWork = FALSE)
}
knitr::opts_knit$set(root.dir = proj_root)
cat("proj_root =", proj_root, "\n")
```

# 0 · What this file does {background-color="#2c3e50"}

## Variant: ρ estimated as a parameter, **AR(2) on growth rate**, FSV

This is the **AR(2) rho-est variant** of the FSV model. Differences
vs the AR(1) rho-est qmd:

1.  **AR(2) on the mean of $\lambda$** with fixed coefficients
    $(\varphi_1, \varphi_2) = (2, -1)$. This matches MARSS structurally:
    MARSS uses $l_t = 2 l_{t-1} - l_{t-2} + w_t$ on the level $l_t$,
    and applying the same shape to the spatial growth rate $\lambda$
    yields direct comparability without adding parameters.
2.  **Two anchors needed**: $\lambda_{p,1}, \lambda_{p,2} \sim
    \mathcal{N}(0, 0.5^2)$ instead of one. Recursion starts at $t = 3$.
3.  **ρ is a PARAMETER**, prior Beta(2, 2) (same as AR(1) rho-est).
4.  **No MARSS Stage 1** — raw `y` with NA mask.
5.  **`log_beta ~ Normal(0, 4)`**, `sigma_obs_pl ~ Half-Normal(0, 2)`
    componentwise — unchanged.
6.  **Stan file**:
    `stan/renewal_A_fsv_est_ar2.stan` (NEW).

**The two knobs that define this experiment:**

``` r
Batch_focus <- paste0("WWTP", 1:15)
ALPHA_RHO   <- 2     # Beta(2, 2) prior shape parameter
BETA_RHO    <- 2     # Beta(2, 2) prior rate parameter
```

**Purpose.** Compare the posterior mean of ρ under joint inference
against the LOO sweep winner. If the two inferences disagree, the
likelihood admits ridge-like flexibility in ρ that LOO selects one
point of but the posterior covers integrally. Also: propagating the
uncertainty in ρ to the predictive widens intervals — expected gain
in empirical 95% coverage of ~10 percentage points relative to fixed
ρ.

# 1 · Packages, paths, helpers {background-color="#2c3e50"}

## Packages

```{r}
#| label: packages
suppressPackageStartupMessages({
  library(tidyverse)
  library(lubridate)
  library(sf)
  # NOTE: MARSS NOT loaded — this variant skips Stage 1.
  library(cmdstanr)      # Stan interface
  library(posterior)     # draws_df, ess, rhat
  library(purrr)
})
set.seed(123)
```

## Settings — the only knobs

```{r}
#| label: settings
# ---- The two knobs that turn this run into "15 plants, rho ESTIMATED" -----
Batch_focus <- paste0("WWTP", 1:15)

# Prior on rho: Beta(ALPHA_RHO, BETA_RHO).
# Defaults to Beta(2, 2): symmetric, mean 0.5, sd 0.224, vanishes at {0, 1}.
# To use a Uniform(0,1) prior, set both to 1.
ALPHA_RHO <- 2
BETA_RHO  <- 2

# rho_grid is kept as a single placeholder so downstream code that references
# rho_grid[1] keeps working — but rho itself is now estimated.
rho_grid  <- c(NA_real_)

# W is built from the population-based distance matrix d_pop_50 only
# (centroid path removed for clarity).

# ---- Input data paths (relative to proj_root for NOTS bundle) --------------
raw_csv_path <- file.path(proj_root, "data",
                          "all_ts_observed_ILI_20250526_pool.csv")
shp_batch    <- file.path(proj_root, "data", "WWTP_Batch",
                          "WWTP_061621_mergedWWTPs to Batch_updatedon07192024.shp")
shp_ct_pop   <- file.path(proj_root, "data", "Census_Tracts",
                          "WWTP Batch Service Areas_ Census Tracts.csv")
csv_ct_pop   <- file.path(proj_root, "data", "Census_Tracts",
                          "WWTP Batch Service Areas_ Census Tracts_Population.csv")

# ---- Tracts shapefile used by the buffer-capture analysis ------------------
tracts_shp_path <- file.path(proj_root, "data", "Census_Tracts_Population",
  "WWTP Batch Service Areas_Census Tracts_Population.shp")

# ---- Pipeline constants ----------------------------------------------------
casos_houston_semana <- 324
burnin               <- 15
cutoff_date_min      <- as.Date("2023-01-02")

prior_mean_gen_days  <- 7.5    # Vink 2014 RSV
prior_sd_gen_days    <- 2.1
mean_shed_days       <- 4.6    # Cevik 2023
sd_shed_days         <- 2.0

# ---- MCMC defaults (Exp B: same as Exp A for direct comparison) -----------
n_chains      <- 3
iter_warmup   <- 2000
iter_sampling <- 8000     # Exp B: 8000 samples → 24000 total post-warmup
adapt_delta   <- 0.95
max_treedepth <- 15
```

## Helpers

```{r}
#| label: helpers
# to_gamma: Gamma shape/scale for a (mean, sd) given in DAYS, returned in WEEKS.
to_gamma <- function(mean_days, sd_days) {
  m <- mean_days / 7; s <- sd_days / 7
  list(shape = (m^2) / (s^2), scale = (s^2) / m)
}

# make_shedding_kernel: discretised Gamma shedding kernel including lag 0.
make_shedding_kernel <- function(mean_shed_days, sd_shed_days, K = 2) {
  par_d   <- to_gamma(mean_shed_days, sd_shed_days)
  edges_d <- c(-0.5, seq(0.5, K + 0.5, by = 1))
  F_d     <- pgamma(pmax(edges_d, 0),
                    shape = par_d$shape, scale = par_d$scale)
  w_d     <- diff(F_d)
  w_d     <- pmax(w_d, 0) / sum(pmax(w_d, 0))
  list(d = w_d, G = K, D = length(w_d))
}
```

# 2 · Shapefile + crosswalk {background-color="#2c3e50"}

## Read the WWTP shapefile and build the WWTP1..WWTP15 crosswalk

```{r}
#| label: shapefile-crosswalk
stopifnot(file.exists(shp_batch))

wwtp_shp_batch <- sf::read_sf(shp_batch) |>
  sf::st_transform(32615) |>
  mutate(BatchName = toupper(trimws(BatchName)))

batch_key_shp <- wwtp_shp_batch |>
  sf::st_drop_geometry() |>
  distinct(BatchName) |>
  arrange(BatchName) |>
  mutate(Batch = paste0("WWTP", row_number()))

wwtp_shp_batch <- wwtp_shp_batch |>
  left_join(batch_key_shp, by = "BatchName")

wwtp_shp_focus <- wwtp_shp_batch |>
  filter(Batch %in% Batch_focus) |>
  arrange(match(Batch, Batch_focus))

stopifnot(nrow(wwtp_shp_focus) == length(Batch_focus))
n_plants <- length(Batch_focus)

cat("Crosswalk WWTPn -> BatchName:\n")
print(batch_key_shp |> filter(Batch %in% Batch_focus))
```

## Census-tract population allocation

```{r}
#| label: ct-allocation
ct_raw <- read_csv(csv_ct_pop, show_col_types = FALSE) |>
  mutate(BatchName = toupper(trimws(BatchName)))

ct_alloc <- ct_raw |>
  left_join(batch_key_shp, by = "BatchName") |>
  filter(!is.na(Batch)) |>
  mutate(
    pop_alloc = Percent_ce * POP24_5YR,
    moe_alloc = Percent_ce * MOE24_5YR
  )

# Geometry for census tracts (prefer real shapefile; fall back to synthetic).
if (file.exists(shp_ct_pop) &&
    grepl("\\.shp$", shp_ct_pop, ignore.case = TRUE)) {
  ct_sf <- sf::read_sf(shp_ct_pop) |>
    sf::st_transform(32615) |>
    mutate(BatchName = toupper(trimws(BatchName))) |>
    left_join(
      ct_alloc |> select(BatchName, Percent_ce, POP24_5YR, pop_alloc, Batch),
      by = c("BatchName", "Percent_ce", "POP24_5YR")
    ) |>
    filter(!is.na(Batch))
} else {
  message("Census tract shapefile not found — generating synthetic centroids.")
  ct_alloc_sf <- ct_alloc |>
    left_join(wwtp_shp_batch |> select(BatchName, geometry), by = "BatchName") |>
    sf::st_as_sf()
  set.seed(999)
  pts <- lapply(seq_len(nrow(ct_alloc_sf)), function(i) {
    s <- tryCatch(sf::st_sample(ct_alloc_sf[i, ], size = 1, type = "random"),
                  error = function(e) NULL)
    if (is.null(s) || length(s) == 0)
      s <- sf::st_point_on_surface(sf::st_geometry(ct_alloc_sf[i, ]))
    s
  })
  ct_sf <- ct_alloc |>
    mutate(geometry = do.call(c, pts)) |>
    sf::st_as_sf(crs = 32615)
}

ct_pop_by_batch <- ct_alloc |>
  group_by(Batch) |>
  summarise(N_acs = sum(pop_alloc, na.rm = TRUE),
            N_moe_acs = sqrt(sum(moe_alloc^2, na.rm = TRUE)),
            .groups = "drop")

cat("\nPopulation per batch:\n"); print(ct_pop_by_batch)
```

# 3 · Wastewater data {background-color="#2c3e50"}

## Read CSV + apply the batch-code crosswalk

```{r}
#| label: wastewater-load
batch_crosswalk <- tibble(
  batch_code = c("B69", "BAS", "BIA", "BKW", "BNO", "BNE", "BNW",
                 "BSB", "BSO", "BSE", "BWE", "BWS", "BKB", "BSW", "BWB"),
  BatchName  = c("69TH STREET", "ALMEDA SIMS", "BATCH INTERCONTINENTAL",
                 "BATCH KINGWOOD", "BATCH NORTH", "BATCH NORTHEAST",
                 "BATCH NORTHWEST", "BATCH SIMS BAYOU", "BATCH SOUTH",
                 "BATCH SOUTHEAST", "BATCH WEST", "BATCH WEST SOUTH",
                 "KEEGANS BAYOU", "SOUTHWEST", "WILLOWBROOK")
)

ww_input <- read_csv(raw_csv_path, show_col_types = FALSE) |>
  mutate(WWTP  = toupper(trimws(WWTP)),
         tn    = toupper(trimws(tn)),
         BATCH = toupper(trimws(BATCH))) |>
  filter(WWTP != "SS", tn == "RSV") |>
  transmute(batch_code = BATCH, WWTP = WWTP, n_plants = Bnum,
            Date = as.Date(date), Pathogen = tn,
            copies_L = copies_L, Flow = flow_interp) |>
  arrange(batch_code, Date)

ww_input <- ww_input |>
  group_by(batch_code) |>
  mutate(median_flow = median(Flow, na.rm = TRUE)) |>
  ungroup() |>
  mutate(copies_day = (copies_L * median_flow) / 1e9) |>
  left_join(batch_crosswalk, by = "batch_code") |>
  left_join(batch_key_shp,   by = "BatchName")

batches_in_data <- ww_input |> filter(!is.na(Batch)) |>
  distinct(Batch) |> pull(Batch)
missing_focus <- setdiff(Batch_focus, batches_in_data)
if (length(missing_focus) > 0)
  stop("These Batch_focus plants have no wastewater data: ",
       paste(missing_focus, collapse = ", "))

ww <- ww_input |>
  filter(Batch %in% Batch_focus) |>
  arrange(Batch, Date)

cat("Plants present in CSV:", length(batches_in_data),
    " — focus uses:", length(Batch_focus), "\n")
```

# 4 · `N_plantas` and `It_init_vec` {background-color="#2c3e50"}

```{r}
#| label: pop-and-It-init
N_plantas <- ct_pop_by_batch |>
  filter(Batch %in% Batch_focus) |>
  arrange(match(Batch, Batch_focus)) |>
  { \(d) setNames(d$N_acs, d$Batch) }()

N_houston <- sum(ct_pop_by_batch$N_acs, na.rm = TRUE)

# Allocate the weekly Houston-wide case count to each plant proportionally
# to its served population, then floor at 10 (Poisson→Lognormal stability).
It_init_vec <- pmax(
  round(casos_houston_semana * N_plantas[Batch_focus] / N_houston),
  10
)
stopifnot(all(It_init_vec >= 10))

cat("N_plantas (1k):\n"); print(round(N_plantas / 1e3, 1))
cat("\nIt_init_vec:\n"); print(It_init_vec)
```

# 5 · `y_batch` — viral load matrix {background-color="#2c3e50"}

```{r}
#| label: y-batch
ww_batch <- ww |>
  select(Batch, n_plants, Date, Pathogen, copies_day, flow_batch = Flow)

cutoff_date <- ww_batch |>
  summarise(d = dplyr::nth(sort(unique(Date)), burnin,
                           default = max(Date))) |>
  pull(d) |>
  max(cutoff_date_min)

ww_rsv_batch <- ww_batch |>
  filter(Date > cutoff_date) |>
  mutate(copies_day = ifelse(is.na(copies_day) | copies_day <= 0,
                              NA_real_, copies_day),
         Missing    = is.na(copies_day)) |>
  arrange(Batch, Date)

make_wide_matrix <- function(df, value_col, plants_order) {
  df |>
    mutate(.val = {{ value_col }}) |>
    select(Date, WWTP = Batch, .val) |>
    dplyr::group_by(Date, WWTP) |>
    dplyr::summarise(.val = if (all(is.na(.val))) NA_real_ else mean(.val, na.rm = TRUE),
                     .groups = "drop") |>
    pivot_wider(names_from = WWTP, values_from = .val, values_fill = NA_real_) |>
    arrange(Date) |>
    { \(w) {
      missing_cols <- setdiff(plants_order, names(w))
      for (m in missing_cols) w[[m]] <- NA_real_
      w |> select(all_of(c("Date", plants_order)))
    } }()
}

y_wide_batch <- make_wide_matrix(ww_rsv_batch, copies_day, Batch_focus)
y_batch      <- as.matrix(y_wide_batch[, -1, drop = FALSE])
rownames(y_batch) <- as.character(y_wide_batch$Date)

stopifnot(identical(colnames(y_batch), Batch_focus))

cat("y_batch:", nrow(y_batch), "weeks x", ncol(y_batch), "plants\n")
cat("Missing cells:", sum(is.na(y_batch)), "/",
    prod(dim(y_batch)), "\n")
```

# 6 · Generation + shedding kernels {background-color="#2c3e50"}

```{r}
#| label: kernels
K_shed <- 2     # shedding kernel support (in weeks beyond lag 0)
ks    <- make_shedding_kernel(mean_shed_days, sd_shed_days, K = K_shed)
d_vec <- as.numeric(ks$d)         # shedding (lag 0..K_shed)
G     <- ks$G                      # generation support length
D     <- ks$D

par_g <- to_gamma(prior_mean_gen_days, prior_sd_gen_days)
edges <- seq(1.5, G + 0.5, by = 1)
g_raw <- pgamma(edges, shape = par_g$shape, scale = par_g$scale) -
         pgamma(edges - 1, shape = par_g$shape, scale = par_g$scale)
g_raw <- pmax(g_raw, 0)
g_vec <- g_raw / sum(g_raw)        # generation kernel (sums to 1)

cat("g_vec (gen kernel, length", G, "):\n");  print(round(g_vec, 4))
cat("d_vec (shed kernel, length", D, "):\n"); print(round(d_vec, 4))
```

# 7 · Buffer-capture analysis on the 15 plants {background-color="#2c3e50"}

## Source helpers + read tracts

```{r}
#| label: source-buffer-capture
source(file.path(proj_root, "R", "core.R"))
source(file.path(proj_root, "R", "plot.R"))
source(file.path(proj_root, "R", "analysis.R"))

tracts_raw <- sf::st_read(tracts_shp_path, quiet = TRUE) |>
  sf::st_make_valid() |>
  dplyr::filter(sf::st_is_valid(geometry), !sf::st_is_empty(geometry))

tracts_valid <- tracts_raw |>
  dplyr::distinct(GEOID, .keep_all = TRUE) |>
  sf::st_transform(32615)

cat("tracts_valid:", nrow(tracts_valid), "unique tracts after dedupe.\n")
```

## Distance matrix: `d_pop_50`

```{r}
#| label: distance-matrices
d_pop_50 <- region_distance_matrix(
  regions         = wwtp_shp_focus,
  label_col       = "BatchName",
  method          = "population",
  tracts          = tracts_valid,
  pop_col         = "POP24_5YR",
  target_fraction = 0.5,
  people_per_dot  = 200,
  units           = "km"
)

cat("d_pop_50: ", paste(dim(d_pop_50), collapse = " x "), "\n", sep = "")
```

## Heatmap of the distance matrix

```{r}
#| label: distance-heatmaps
#| fig-width: 8
#| fig-height: 7
#| fig-show: "hide"
#| results: "hide"
dir.create("outputs/plots", showWarnings = FALSE, recursive = TRUE)
pdf("outputs/plots/distance_heatmap_pop50.pdf", width = 8, height = 7)
plot_distance_heatmap(d_pop_50, cluster = TRUE)
invisible(dev.off())
cat("→ outputs/plots/distance_heatmap_pop50.pdf\n")
```

# 8 · Distance → spatial weight matrix W {background-color="#2c3e50"}

```{r}
#| label: d-to-W
make_W_from_distance <- function(d, sigma = NULL) {
  stopifnot(nrow(d) == ncol(d))
  d_dim   <- dim(d)
  d_names <- dimnames(d)
  d <- structure(as.numeric(unclass(d)),
                 dim = d_dim, dimnames = d_names)

  off <- d
  diag(off) <- NA
  if (is.null(sigma)) {
    # Median heuristic on SQUARED distances (Gretton et al. 2012).
    sigma <- sqrt(median(off^2, na.rm = TRUE) / 2)
  }

  K <- exp(-d^2 / (2 * sigma^2))
  diag(K) <- 0
  rs <- rowSums(K)
  if (any(rs == 0))
    stop("Row sum of K is zero — pick a larger sigma.")
  W <- K / rs
  attr(W, "sigma_km") <- sigma
  attr(W, "kernel")   <- "gaussian"
  W
}

W_pop50 <- make_W_from_distance(d_pop_50)

cat(sprintf("kernel: %s, sigma (km): %.2f\n",
            attr(W_pop50, "kernel"),
            attr(W_pop50, "sigma_km")))

stopifnot(all(abs(rowSums(W_pop50) - 1) < 1e-10),
          all(diag(W_pop50) == 0))
```

## Rename BatchName → WWTPn and align with `Batch_focus`

```{r}
#| label: rename-W-WWTP
bn_to_wwtp <- setNames(batch_key_shp$Batch, batch_key_shp$BatchName)
stopifnot(all(rownames(W_pop50) %in% names(bn_to_wwtp)))

rename_to_wwtp <- function(W) {
  rownames(W) <- bn_to_wwtp[rownames(W)]
  colnames(W) <- bn_to_wwtp[colnames(W)]
  W[Batch_focus, Batch_focus, drop = FALSE]
}
W_pop50 <- rename_to_wwtp(W_pop50)

# The W that goes into Stan.
W_geo <- W_pop50

stopifnot(identical(rownames(W_geo), Batch_focus),
          identical(colnames(W_geo), Batch_focus),
          identical(rownames(W_geo), colnames(y_batch)))

cat(sprintf("W_geo: %d x %d (source = pop50)\n",
            nrow(W_geo), ncol(W_geo)))
```

# 9 · Stan model (freesigma_vec variant) {background-color="#2c3e50"}

## Read the `.stan` file from disk

We **read** the freesigma_vec Stan file rather than write it inline.
The source of truth for this model is
`stan/renewal_A_fsv.stan`, which is shared with the NOTS
script `run_stan_freesigma_vec.R`. This prevents drift between the qmd
and the .R variants.

Key model features:

-   `log_beta ~ Normal(0, 4)` — weakly informative, \~5 orders of
    magnitude on β.
-   `sigma_obs_pl[p] ~ Half-Normal(0, 2)` (componentwise) — Gelman
    (2006) prior for variance components.
-   `sigma_obs_pl` is a **vector\[P\]**, one σ per plant. Cambio
    principal vs el modelo scalar de freesigma.
-   No `sigma_obs` data input, no `use_filter` flag.

```{r}
#| label: read-stan-model
stan_path <- file.path(proj_root, "stan",
                       "renewal_A_fsv_est_ar2.stan")
stopifnot(file.exists(stan_path))

cat("Using Stan model:\n  ", stan_path, "\n", sep = "")

# Sanity preview: first 60 lines
readLines(stan_path) |> head(60) |> writeLines()
```

# 10 · Build the Stan data list {background-color="#2c3e50"}

```{r}
#| label: build-stan-data
# y in P x T (Stan convention); existing y_batch is T x P.
y_PT <- t(y_batch)
stopifnot(identical(rownames(y_PT), Batch_focus),
          identical(rownames(W_geo), Batch_focus))

# Mask + dummy fill for NA cells (Stan rejects NA in `data`).
y_obs_mat <- matrix(as.integer(is.finite(y_PT) & y_PT > 0),
                    nrow = nrow(y_PT), ncol = ncol(y_PT))
y_filled  <- y_PT
y_filled[y_obs_mat == 0] <- 1.0

P <- length(Batch_focus)
T <- ncol(y_PT)

# NOTA rho-est: rho ya NO es data. Es parametro con prior Beta(alpha_rho, beta_rho).
# sigma_obs_pl sigue siendo parametro libre vector[P].
build_stan_data <- function(...) {
  list(
    P            = P,
    T            = T,
    G            = G,
    D            = D,
    g            = as.numeric(g_vec),
    d            = as.numeric(d_vec),
    W            = W_geo,
    # rho REMOVED — now estimated as parameter
    It_init      = as.numeric(It_init_vec),
    y            = y_filled,
    y_obs        = y_obs_mat,
    mean_gen_wk  = prior_mean_gen_days / 7,
    sd_gen_wk    = prior_sd_gen_days   / 7,
    # NEW: hyperparameters for Beta(alpha_rho, beta_rho) prior on rho
    alpha_rho    = ALPHA_RHO,
    beta_rho     = BETA_RHO
  )
}

stan_data_demo <- build_stan_data()
str(stan_data_demo, max.level = 1)
cat(sprintf("\nPrior on rho: Beta(%g, %g)  -> mean = %.3f, sd = %.3f\n",
            ALPHA_RHO, BETA_RHO,
            ALPHA_RHO / (ALPHA_RHO + BETA_RHO),
            sqrt(ALPHA_RHO * BETA_RHO /
                 ((ALPHA_RHO + BETA_RHO)^2 * (ALPHA_RHO + BETA_RHO + 1)))))
```

# 11 · Compile Stan model {background-color="#2c3e50"}

```{r}
#| label: compile-stan
## stan_threads = TRUE matches the cached binary at stan/rsv_renewal_freesigma_vec,
## so cmdstanr reuses it instead of recompiling. Edit the .stan content and
## run `cmdstanr::rebuild_cmdstan()` once to refresh the PCH if needed.
mod <- cmdstan_model(stan_file = stan_path, compile = TRUE,
                     cpp_options = list(stan_threads = TRUE))

cat("Compiled binary at:", mod$exe_file(), "\n")
```

# 12 · Single fit — rho estimated jointly {background-color="#2c3e50"}

## The single-fit run

Unlike the sweep version, this experiment fits the model **once** with
rho as a parameter (Beta(`ALPHA_RHO`, `BETA_RHO`) prior). The fit is
persisted to `outputs/stan_fits_fsv_rhoest/rho_estimated.rds`.

```{r}
#| label: stan-sample-rhoest
#| eval: true

RUN_SAMPLE <- TRUE

fits_dir <- file.path(proj_root, "outputs", "stan_fits_fsv_rhoest_ar2")
dir.create(fits_dir, showWarnings = FALSE, recursive = TRUE)

# init: sigma_obs_pl VECTOR[P], rho centrado en 0.5 (prior mean Beta(2,2)).
init_fun <- function(chain_id) {
  list(
    log_beta     = rep(0, P),
    sigma_obs_pl = rep(1.0, P),
    lam          = matrix(0.1, P, T),
    It           = matrix(rep(It_init_vec, T), P, T),
    rho          = 0.5   # arrancar en el centro del prior
  )
}
init_list <- lapply(seq_len(n_chains), init_fun)

if (RUN_SAMPLE) {
  cat(sprintf("Sampling ONCE with rho as parameter (Beta(%g, %g) prior)\n",
              ALPHA_RHO, BETA_RHO))
  t0 <- Sys.time()

  sd_one <- build_stan_data()
  fit <- mod$sample(
    data              = sd_one,
    init              = init_list,
    chains            = n_chains,
    parallel_chains   = n_chains,
    threads_per_chain = 1,
    iter_warmup       = iter_warmup,
    iter_sampling     = iter_sampling,
    adapt_delta       = adapt_delta,
    max_treedepth     = max_treedepth,
    refresh           = 500,
    seed              = 1729
  )

  cat(sprintf("\nSampling done in %.1f min\n",
              as.numeric(difftime(Sys.time(), t0, units = "mins"))))

  cat("Extracting draws (includes rho) ...\n")
  draws_arr <- fit$draws(
    variables = c("log_beta", "beta", "sigma_obs_pl", "Rt", "It",
                  "rho", "log_lik", "lp__"),
    format = "draws_array")
  diag_sum  <- fit$diagnostic_summary()
  summary_tbl <- fit$summary(
    variables = c("log_beta", "beta", "sigma_obs_pl", "Rt", "It",
                  "rho", "lp__"),
    rhat = posterior::rhat,
    ess_bulk = posterior::ess_bulk,
    ess_tail = posterior::ess_tail,
    mean, sd,
    q5  = ~ quantile(.x, 0.05),
    q50 = ~ quantile(.x, 0.50),
    q95 = ~ quantile(.x, 0.95)
  )

  # rho posterior summary (prints front and center)
  rho_post <- summary_tbl[summary_tbl$variable == "rho", ]
  cat("\n=== POSTERIOR OF RHO ===\n")
  print(rho_post)

  save_object <- list(
    rho_prior             = list(family = "beta",
                                 alpha  = ALPHA_RHO,
                                 beta   = BETA_RHO),
    rho_posterior_summary = rho_post,
    model_variant         = "freesigma_vec_rhoest",
    draws                 = draws_arr,
    diagnostic_summary    = diag_sum,
    summary               = summary_tbl,
    y_raw                 = y_PT,
    y_obs_mask            = y_obs_mat,
    d_vec                 = d_vec,
    meta = list(
      n_chains        = n_chains,
      iter_warmup     = iter_warmup,
      iter_sampling   = iter_sampling,
      adapt_delta     = adapt_delta,
      max_treedepth   = max_treedepth,
      log_beta_prior  = "N(0, 4)",
      sigma_obs_prior = "Half-Normal(0, 2) per plant",
      rho_prior       = sprintf("Beta(%g, %g)", ALPHA_RHO, BETA_RHO),
      save_time       = Sys.time()
    )
  )

  rds_path <- file.path(fits_dir, "rho_estimated.rds")
  tmp_path <- file.path(Sys.getenv("TMPDIR", tempdir()),
                        sprintf("rho_estimated_%d.rds", Sys.getpid()))
  saveRDS(save_object, tmp_path, compress = "gzip")
  ok <- file.copy(tmp_path, rds_path, overwrite = TRUE)
  if (!ok) stop("file.copy failed")
  file.remove(tmp_path)
  cat(sprintf("Saved fit -> %s  (%.1f MB)\n",
              rds_path, file.size(rds_path) / 1e6))

  fits <- list(rho_estimated = save_object)
} else {
  rds_path <- file.path(fits_dir, "rho_estimated.rds")
  if (file.exists(rds_path)) {
    fits <- list(rho_estimated = readRDS(rds_path))
    cat("Loaded existing fit:", rds_path, "\n")
  } else {
    stop("No existing fit and RUN_SAMPLE = FALSE.")
  }
}

# 13 · Convergence + rho posterior diagnostics {background-color="#2c3e50"}

## R-hat, ESS, divergencias + posterior summary de ρ

```{r}
#| label: diagnostics-rhoest
#| eval: true
if (RUN_SAMPLE || exists("fits")) {
  fit_obj <- fits$rho_estimated
  sm <- fit_obj$summary
  # Resumen global de convergencia (todos los params)
  diag_global <- data.frame(
    n_divergent       = sum(fit_obj$diagnostic_summary$num_divergent),
    n_max_treedepth   = sum(fit_obj$diagnostic_summary$num_max_treedepth),
    rhat_max          = max(sm$rhat, na.rm = TRUE),
    pct_rhat_above_1_01 = round(100 * mean(sm$rhat > 1.01, na.rm = TRUE), 2),
    ess_bulk_min      = min(sm$ess_bulk, na.rm = TRUE),
    pct_ess_below_400 = round(100 * mean(sm$ess_bulk < 400, na.rm = TRUE), 2)
  )
  knitr::kable(diag_global,
               caption = "Convergence summary (all parameters).")

  # Posterior de rho
  rho_row <- sm[sm$variable == "rho", ]
  knitr::kable(rho_row,
               digits = 4,
               caption = sprintf("Posterior of rho under Beta(%g, %g) prior.",
                                 ALPHA_RHO, BETA_RHO))

  # sigma_obs_pl posterior (per plant)
  sigma_rows <- sm[grepl("^sigma_obs_pl", sm$variable), ]
  cat(sprintf("\nsigma_obs_pl across %d plants: median = %.3f, range = [%.3f, %.3f]\n",
              nrow(sigma_rows),
              median(sigma_rows$mean),
              min(sigma_rows$mean),
              max(sigma_rows$mean)))
}
```

## Histogram of rho posterior draws

```{r}
#| label: rho-posterior-histogram
#| eval: true
#| fig-height: 4
if (RUN_SAMPLE || exists("fits")) {
  rho_draws <- as.numeric(posterior::as_draws_matrix(
    posterior::subset_draws(fits$rho_estimated$draws, variable = "rho")))
  hist(rho_draws, breaks = 50, freq = FALSE,
       main = bquote("Posterior of " * rho ~
                     " | Beta(" * .(ALPHA_RHO) * "," * .(BETA_RHO) * ") prior"),
       xlab = expression(rho),
       col = "#1d4ed8", border = "white", xlim = c(0, 1))
  curve(dbeta(x, ALPHA_RHO, BETA_RHO), add = TRUE,
        col = "grey50", lty = 2, lwd = 2)
  abline(v = mean(rho_draws), col = "red", lwd = 2)
  legend("topright",
         legend = c("Posterior", "Prior Beta(2,2)",
                    sprintf("Mean = %.3f", mean(rho_draws))),
         fill = c("#1d4ed8", NA, NA),
         border = c("white", NA, NA),
         lty = c(NA, 2, 1),
         col = c(NA, "grey50", "red"),
         lwd = c(NA, 2, 2),
         bty = "n", cex = 0.9)
}
```

# 14 · Notes and known limitations {background-color="#2c3e50"}

-   **rho is now estimated**, not fixed. Posterior summary captures
    both the central tendency (mean / median) and the data-driven
    uncertainty (sd, 95% CrI). Expect a tight posterior far from the
    prior mean if ρ is well-identified by the data.
-   **No MARSS Stage 1.** Same as FSV sweep: raw `y` with NA mask, and
    `sigma_obs_pl[p]` as a free vector parameter (Half-Normal(0, 2)
    componentwise).
-   **Comparison to sweep.** With the sweep, fitting was done at each
    fixed ρ in `{0.0, 0.1, ..., 0.9}` independently. Here a SINGLE fit
    samples ρ jointly. Use `analyze_rhoest.R` and
    `analyze_contrast_rho0.R` (in the project root) to compare the
    rho-est posterior against the sweep's ELPD-LOO winner.
-   **Wall-clock cost.** Single fit with 3 chains × 2000 warmup ×
    8000 sampling and P = 15: ~1.5–3× slower than a fixed-ρ sweep fit
    because ρ now mixes with all the other parameters through `I_mix`
    and `mu_lam`. Expect ~3–6 hours on a 20-core node.
-   **Output dir.** `.rds` lands at
    `outputs/stan_fits_fsv_rhoest/rho_estimated.rds` (single file,
    not a per-ρ grid). The companion MARSS rho-est variant uses
    `stan/renewal_A_marss_est.stan` and saves to
    `outputs/stan_fits_marss_rhoest/`.
