---
title: "Spatial RSV Renewal — Stan (freesigma_vec, σ por planta)"
subtitle: "Self-contained: data prep + buffer-capture W + rho sweep, vector[P] σ"
author: "Jose Palacio"
date: today
date-format: "MMMM D, YYYY"
format:
  revealjs:
    theme: simple
    slide-number: true
    progress: true
    code-line-numbers: true
    highlight-style: github
    footer: "Spatial RSV Renewal · Stan · 15 plants · freesigma_vec"
    transition: fade
    background-transition: fade
    incremental: false
    toc: false
    smaller: true
    scrollable: true
    fig-align: center
    fig-width: 9
    fig-height: 5
execute:
  echo: true
  eval: true
  warning: false
  message: false
  cache: false
knitr:
  opts_chunk:
    comment: "#>"
editor_options:
  chunk_output_type: console
editor:
  markdown:
    wrap: 72
---

```{=html}
<style>
:root { --font-scale: 0.75; --code-scale: 0.75; }
.reveal .slides         { font-size: calc(40px * var(--font-scale)); }
.reveal .slides h1      { font-size: calc(2.2em * var(--font-scale)); }
.reveal .slides h2      { font-size: calc(1.6em * var(--font-scale)); }
.reveal pre, .reveal pre code { font-size: calc(0.75em * var(--code-scale)); }
.reveal table           { font-size: calc(0.8em * var(--font-scale)); }
</style>
```

```{r}
#| label: root-dir
#| include: false
# proj_root: ruta absoluta a la carpeta donde viven data/, stan/, R/.
# Local Mac: usa absoluta para que funcione desde cualquier cwd.
# NOTS: cambiar a `normalizePath(".")` (cwd del SLURM submit dir).
proj_root <- if (Sys.info()["sysname"] == "Darwin") {
  "/Users/josepalacio/Documents/Claude/Projects/NOTS/rsv_mcmc_stan_geoW"
} else {
  normalizePath(".", mustWork = FALSE)
}
knitr::opts_knit$set(root.dir = proj_root)
cat("proj_root =", proj_root, "\n")
```

# 0 · What this file does {background-color="#2c3e50"}

## Variant: σ_obs por planta (vector[P]), NO MARSS

This is the **freesigma_vec variant** of the main Stan model.
Differences vs `SPSTAN_GR_IN_15plants_geoW-freesigma.qmd` (the scalar
version):

1.  **No MARSS Stage 1.** Same as freesigma — no `sigma_filt`, no
    `use_filter` flag.
2.  **σ_obs is a VECTOR[P] of FREE parameters.** Each plant has its own
    `sigma_obs_pl[p]`, with prior
    `sigma_obs_pl[p] ~ Half-Normal(0, 2)` componentwise. This addresses
    the **coverage collapse** observed with the scalar version
    (empirical 50% at nominal 95%, because a single σ cannot capture
    heterogeneity between WWTP15 noise=0.98 and WWTP14 noise=0.47).
3.  **`log_beta ~ Normal(0, 4)`** — same wide prior. Expectation: β CIs
    will WIDEN with this fix because σ now absorbs more variance.
4.  **Stan file**: `stan/renewal_A_fsv.stan` (NOT
    `rsv_renewal_freesigma.stan`).

**The two knobs you will change most:**

``` r
Batch_focus <- paste0("WWTP", 1:15)
rho_grid    <- c(0.0)            # ρ = 0 baseline (independent-plant fit)
```

**Purpose of this experiment.** Validate that vectorizing σ_obs fixes
the coverage. If empirical 95% coverage jumps from ~50% (scalar
freesigma) to ~85-95%, then the heterogeneity between plants was the
main miscalibration source.

# 1 · Packages, paths, helpers {background-color="#2c3e50"}

## Packages

```{r}
#| label: packages
suppressPackageStartupMessages({
  library(tidyverse)
  library(lubridate)
  library(sf)
  # NOTE: MARSS NOT loaded — this variant skips Stage 1.
  library(cmdstanr)      # Stan interface
  library(posterior)     # draws_df, ess, rhat
  library(purrr)
})
set.seed(123)
```

## Settings — the only knobs

```{r}
#| label: settings
# ---- The two knobs for this run: 15 plants, ρ = 0 baseline ----------------
Batch_focus <- paste0("WWTP", 1:15)
rho_grid    <- c(0.0)   # Independent-plant fit (no spatial coupling active).

# W is built from the population-based distance matrix d_pop_50 only
# (centroid path removed for clarity).

# ---- Input data paths (relative to proj_root for NOTS bundle) --------------
raw_csv_path <- file.path(proj_root, "data",
                          "all_ts_observed_ILI_20250526_pool.csv")
shp_batch    <- file.path(proj_root, "data", "WWTP_Batch",
                          "WWTP_061621_mergedWWTPs to Batch_updatedon07192024.shp")
shp_ct_pop   <- file.path(proj_root, "data", "Census_Tracts",
                          "WWTP Batch Service Areas_ Census Tracts.csv")
csv_ct_pop   <- file.path(proj_root, "data", "Census_Tracts",
                          "WWTP Batch Service Areas_ Census Tracts_Population.csv")

# ---- Tracts shapefile used by the buffer-capture analysis ------------------
tracts_shp_path <- file.path(proj_root, "data", "Census_Tracts_Population",
  "WWTP Batch Service Areas_Census Tracts_Population.shp")

# ---- Pipeline constants ----------------------------------------------------
casos_houston_semana <- 324
burnin               <- 15
cutoff_date_min      <- as.Date("2023-01-02")

prior_mean_gen_days  <- 7.5    # Vink 2014 RSV
prior_sd_gen_days    <- 2.1
mean_shed_days       <- 4.6    # Cevik 2023
sd_shed_days         <- 2.0

# ---- MCMC defaults (Exp B: same as Exp A for direct comparison) -----------
n_chains      <- 3
iter_warmup   <- 2000
iter_sampling <- 8000     
adapt_delta   <- 0.95
max_treedepth <- 15
```

## Helpers

```{r}
#| label: helpers
# to_gamma: Gamma shape/scale for a (mean, sd) given in DAYS, returned in WEEKS.
to_gamma <- function(mean_days, sd_days) {
  m <- mean_days / 7; s <- sd_days / 7
  list(shape = (m^2) / (s^2), scale = (s^2) / m)
}

# make_shedding_kernel: discretised Gamma shedding kernel including lag 0.
make_shedding_kernel <- function(mean_shed_days, sd_shed_days, K = 2) {
  par_d   <- to_gamma(mean_shed_days, sd_shed_days)
  edges_d <- c(-0.5, seq(0.5, K + 0.5, by = 1))
  F_d     <- pgamma(pmax(edges_d, 0),
                    shape = par_d$shape, scale = par_d$scale)
  w_d     <- diff(F_d)
  w_d     <- pmax(w_d, 0) / sum(pmax(w_d, 0))
  list(d = w_d, G = K, D = length(w_d))
}
```

# 2 · Shapefile + crosswalk {background-color="#2c3e50"}

*Paper §2.1 — Data and study area (population allocation via eq. 2).*

## Read the WWTP shapefile and build the WWTP1..WWTP15 crosswalk

```{r}
#| label: shapefile-crosswalk
stopifnot(file.exists(shp_batch))

wwtp_shp_batch <- sf::read_sf(shp_batch) |>
  sf::st_transform(32615) |>
  mutate(BatchName = toupper(trimws(BatchName)))

batch_key_shp <- wwtp_shp_batch |>
  sf::st_drop_geometry() |>
  distinct(BatchName) |>
  arrange(BatchName) |>
  mutate(Batch = paste0("WWTP", row_number()))

wwtp_shp_batch <- wwtp_shp_batch |>
  left_join(batch_key_shp, by = "BatchName")

wwtp_shp_focus <- wwtp_shp_batch |>
  filter(Batch %in% Batch_focus) |>
  arrange(match(Batch, Batch_focus))

stopifnot(nrow(wwtp_shp_focus) == length(Batch_focus))
n_plants <- length(Batch_focus)

cat("Crosswalk WWTPn -> BatchName:\n")
print(batch_key_shp |> filter(Batch %in% Batch_focus))
```

## Census-tract population allocation

```{r}
#| label: ct-allocation
ct_raw <- read_csv(csv_ct_pop, show_col_types = FALSE) |>
  mutate(BatchName = toupper(trimws(BatchName)))

ct_alloc <- ct_raw |>
  left_join(batch_key_shp, by = "BatchName") |>
  filter(!is.na(Batch)) |>
  mutate(
    pop_alloc = Percent_ce * POP24_5YR,
    moe_alloc = Percent_ce * MOE24_5YR
  )

# Geometry for census tracts (prefer real shapefile; fall back to synthetic).
if (file.exists(shp_ct_pop) &&
    grepl("\\.shp$", shp_ct_pop, ignore.case = TRUE)) {
  ct_sf <- sf::read_sf(shp_ct_pop) |>
    sf::st_transform(32615) |>
    mutate(BatchName = toupper(trimws(BatchName))) |>
    left_join(
      ct_alloc |> select(BatchName, Percent_ce, POP24_5YR, pop_alloc, Batch),
      by = c("BatchName", "Percent_ce", "POP24_5YR")
    ) |>
    filter(!is.na(Batch))
} else {
  message("Census tract shapefile not found — generating synthetic centroids.")
  ct_alloc_sf <- ct_alloc |>
    left_join(wwtp_shp_batch |> select(BatchName, geometry), by = "BatchName") |>
    sf::st_as_sf()
  set.seed(999)
  pts <- lapply(seq_len(nrow(ct_alloc_sf)), function(i) {
    s <- tryCatch(sf::st_sample(ct_alloc_sf[i, ], size = 1, type = "random"),
                  error = function(e) NULL)
    if (is.null(s) || length(s) == 0)
      s <- sf::st_point_on_surface(sf::st_geometry(ct_alloc_sf[i, ]))
    s
  })
  ct_sf <- ct_alloc |>
    mutate(geometry = do.call(c, pts)) |>
    sf::st_as_sf(crs = 32615)
}

ct_pop_by_batch <- ct_alloc |>
  group_by(Batch) |>
  summarise(N_acs = sum(pop_alloc, na.rm = TRUE),
            N_moe_acs = sqrt(sum(moe_alloc^2, na.rm = TRUE)),
            .groups = "drop")

cat("\nPopulation per batch:\n"); print(ct_pop_by_batch)
```

# 3 · Wastewater data {background-color="#2c3e50"}

*Paper §2.1 — viral load matrix (eq. 1).*

## Read CSV + apply the batch-code crosswalk

```{r}
#| label: wastewater-load
batch_crosswalk <- tibble(
  batch_code = c("B69", "BAS", "BIA", "BKW", "BNO", "BNE", "BNW",
                 "BSB", "BSO", "BSE", "BWE", "BWS", "BKB", "BSW", "BWB"),
  BatchName  = c("69TH STREET", "ALMEDA SIMS", "BATCH INTERCONTINENTAL",
                 "BATCH KINGWOOD", "BATCH NORTH", "BATCH NORTHEAST",
                 "BATCH NORTHWEST", "BATCH SIMS BAYOU", "BATCH SOUTH",
                 "BATCH SOUTHEAST", "BATCH WEST", "BATCH WEST SOUTH",
                 "KEEGANS BAYOU", "SOUTHWEST", "WILLOWBROOK")
)

ww_input <- read_csv(raw_csv_path, show_col_types = FALSE) |>
  mutate(WWTP  = toupper(trimws(WWTP)),
         tn    = toupper(trimws(tn)),
         BATCH = toupper(trimws(BATCH))) |>
  filter(WWTP != "SS", tn == "RSV") |>
  transmute(batch_code = BATCH, WWTP = WWTP, n_plants = Bnum,
            Date = as.Date(date), Pathogen = tn,
            copies_L = copies_L, Flow = flow_interp) |>
  arrange(batch_code, Date)

ww_input <- ww_input |>
  group_by(batch_code) |>
  mutate(median_flow = median(Flow, na.rm = TRUE)) |>
  ungroup() |>
  mutate(copies_day = (copies_L * median_flow) / 1e9) |>
  left_join(batch_crosswalk, by = "batch_code") |>
  left_join(batch_key_shp,   by = "BatchName")

batches_in_data <- ww_input |> filter(!is.na(Batch)) |>
  distinct(Batch) |> pull(Batch)
missing_focus <- setdiff(Batch_focus, batches_in_data)
if (length(missing_focus) > 0)
  stop("These Batch_focus plants have no wastewater data: ",
       paste(missing_focus, collapse = ", "))

ww <- ww_input |>
  filter(Batch %in% Batch_focus) |>
  arrange(Batch, Date)

cat("Plants present in CSV:", length(batches_in_data),
    " — focus uses:", length(Batch_focus), "\n")
```

# 4 · `N_plantas` and `It_init_vec` {background-color="#2c3e50"}

*Paper §2.1 (populations, eq. 2) and §2.4.4 (initial incidence anchor, eq. 17).*

```{r}
#| label: pop-and-It-init
N_plantas <- ct_pop_by_batch |>
  filter(Batch %in% Batch_focus) |>
  arrange(match(Batch, Batch_focus)) |>
  { \(d) setNames(d$N_acs, d$Batch) }()

N_houston <- sum(ct_pop_by_batch$N_acs, na.rm = TRUE)

# Allocate the weekly Houston-wide case count to each plant proportionally
# to its served population, then floor at 10 (Poisson→Lognormal stability).
It_init_vec <- pmax(
  round(casos_houston_semana * N_plantas[Batch_focus] / N_houston),
  10
)
stopifnot(all(It_init_vec >= 10))

cat("N_plantas (1k):\n"); print(round(N_plantas / 1e3, 1))
cat("\nIt_init_vec:\n"); print(It_init_vec)
```

# 5 · `y_batch` — viral load matrix {background-color="#2c3e50"}

*Paper §2.1 — plant × week viral load matrix (eq. 1).*

```{r}
#| label: y-batch
ww_batch <- ww |>
  select(Batch, n_plants, Date, Pathogen, copies_day, flow_batch = Flow)

cutoff_date <- ww_batch |>
  summarise(d = dplyr::nth(sort(unique(Date)), burnin,
                           default = max(Date))) |>
  pull(d) |>
  max(cutoff_date_min)

ww_rsv_batch <- ww_batch |>
  filter(Date > cutoff_date) |>
  mutate(copies_day = ifelse(is.na(copies_day) | copies_day <= 0,
                              NA_real_, copies_day),
         Missing    = is.na(copies_day)) |>
  arrange(Batch, Date)

make_wide_matrix <- function(df, value_col, plants_order) {
  df |>
    mutate(.val = {{ value_col }}) |>
    select(Date, WWTP = Batch, .val) |>
    dplyr::group_by(Date, WWTP) |>
    dplyr::summarise(.val = if (all(is.na(.val))) NA_real_ else mean(.val, na.rm = TRUE),
                     .groups = "drop") |>
    pivot_wider(names_from = WWTP, values_from = .val, values_fill = NA_real_) |>
    arrange(Date) |>
    { \(w) {
      missing_cols <- setdiff(plants_order, names(w))
      for (m in missing_cols) w[[m]] <- NA_real_
      w |> select(all_of(c("Date", plants_order)))
    } }()
}

y_wide_batch <- make_wide_matrix(ww_rsv_batch, copies_day, Batch_focus)
y_batch      <- as.matrix(y_wide_batch[, -1, drop = FALSE])
rownames(y_batch) <- as.character(y_wide_batch$Date)

stopifnot(identical(colnames(y_batch), Batch_focus))

cat("y_batch:", nrow(y_batch), "weeks x", ncol(y_batch), "plants\n")
cat("Missing cells:", sum(is.na(y_batch)), "/",
    prod(dim(y_batch)), "\n")
```

# 6 · Generation + shedding kernels {background-color="#2c3e50"}

*Paper §2.3 — generation and shedding kernels (eqs. 7–8).*

```{r}
#| label: kernels
K_shed <- 2     # shedding kernel support (in weeks beyond lag 0)
ks    <- make_shedding_kernel(mean_shed_days, sd_shed_days, K = K_shed)
d_vec <- as.numeric(ks$d)         # shedding (lag 0..K_shed)
G     <- ks$G                      # generation support length
D     <- ks$D

par_g <- to_gamma(prior_mean_gen_days, prior_sd_gen_days)
edges <- seq(1.5, G + 0.5, by = 1)
g_raw <- pgamma(edges, shape = par_g$shape, scale = par_g$scale) -
         pgamma(edges - 1, shape = par_g$shape, scale = par_g$scale)
g_raw <- pmax(g_raw, 0)
g_vec <- g_raw / sum(g_raw)        # generation kernel (sums to 1)

cat("g_vec (gen kernel, length", G, "):\n");  print(round(g_vec, 4))
cat("d_vec (shed kernel, length", D, "):\n"); print(round(d_vec, 4))
```

# 7 · Buffer-capture analysis on the 15 plants {background-color="#2c3e50"}

*Paper §2.2.1 — Population Extended Hausdorff Distance (eqs. 3–4).*

## Source helpers + read tracts

```{r}
#| label: source-buffer-capture
source(file.path(proj_root, "R", "core.R"))
source(file.path(proj_root, "R", "plot.R"))
source(file.path(proj_root, "R", "analysis.R"))

tracts_raw <- sf::st_read(tracts_shp_path, quiet = TRUE) |>
  sf::st_make_valid() |>
  dplyr::filter(sf::st_is_valid(geometry), !sf::st_is_empty(geometry))

tracts_valid <- tracts_raw |>
  dplyr::distinct(GEOID, .keep_all = TRUE) |>
  sf::st_transform(32615)

cat("tracts_valid:", nrow(tracts_valid), "unique tracts after dedupe.\n")
```

## Distance matrix: `d_pop_50`

```{r}
#| label: distance-matrices
d_pop_50 <- region_distance_matrix(
  regions         = wwtp_shp_focus,
  label_col       = "BatchName",
  method          = "population",
  tracts          = tracts_valid,
  pop_col         = "POP24_5YR",
  target_fraction = 0.5,
  people_per_dot  = 200,
  units           = "km"
)

cat("d_pop_50: ", paste(dim(d_pop_50), collapse = " x "), "\n", sep = "")
```

## Heatmap of the distance matrix

```{r}
#| label: distance-heatmaps
#| fig-width: 8
#| fig-height: 7
#| fig-show: "hide"
#| results: "hide"
dir.create("outputs/plots", showWarnings = FALSE, recursive = TRUE)
pdf("outputs/plots/distance_heatmap_pop50.pdf", width = 8, height = 7)
plot_distance_heatmap(d_pop_50, cluster = TRUE)
invisible(dev.off())
cat("→ outputs/plots/distance_heatmap_pop50.pdf\n")
```

# 8 · Distance → spatial weight matrix W {background-color="#2c3e50"}

*Paper §2.2.2 — RBF kernel and row normalization (eqs. 5–6).*

```{r}
#| label: d-to-W
make_W_from_distance <- function(d, sigma = NULL) {
  stopifnot(nrow(d) == ncol(d))
  d_dim   <- dim(d)
  d_names <- dimnames(d)
  d <- structure(as.numeric(unclass(d)),
                 dim = d_dim, dimnames = d_names)

  off <- d
  diag(off) <- NA
  if (is.null(sigma)) {
    # Median heuristic on SQUARED distances (Gretton et al. 2012).
    sigma <- sqrt(median(off^2, na.rm = TRUE) / 2)
  }

  K <- exp(-d^2 / (2 * sigma^2))
  diag(K) <- 0
  rs <- rowSums(K)
  if (any(rs == 0))
    stop("Row sum of K is zero — pick a larger sigma.")
  W <- K / rs
  attr(W, "sigma_km") <- sigma
  attr(W, "kernel")   <- "gaussian"
  W
}

W_pop50 <- make_W_from_distance(d_pop_50)

cat(sprintf("kernel: %s, sigma (km): %.2f\n",
            attr(W_pop50, "kernel"),
            attr(W_pop50, "sigma_km")))

stopifnot(all(abs(rowSums(W_pop50) - 1) < 1e-10),
          all(diag(W_pop50) == 0))
```

## Rename BatchName → WWTPn and align with `Batch_focus`

```{r}
#| label: rename-W-WWTP
bn_to_wwtp <- setNames(batch_key_shp$Batch, batch_key_shp$BatchName)
stopifnot(all(rownames(W_pop50) %in% names(bn_to_wwtp)))

rename_to_wwtp <- function(W) {
  rownames(W) <- bn_to_wwtp[rownames(W)]
  colnames(W) <- bn_to_wwtp[colnames(W)]
  W[Batch_focus, Batch_focus, drop = FALSE]
}
W_pop50 <- rename_to_wwtp(W_pop50)

# The W that goes into Stan.
W_geo <- W_pop50

stopifnot(identical(rownames(W_geo), Batch_focus),
          identical(colnames(W_geo), Batch_focus),
          identical(rownames(W_geo), colnames(y_batch)))

cat(sprintf("W_geo: %d x %d (source = pop50)\n",
            nrow(W_geo), ncol(W_geo)))
```

# 9 · Stan model (freesigma_vec variant) {background-color="#2c3e50"}

*Paper §2.4 (Model specification) and §2.5 (Priors and posterior inference).*

## Read the `.stan` file from disk

We **read** the freesigma_vec Stan file rather than write it inline.
The source of truth for this model is
`stan/renewal_A_fsv.stan`, which is shared with the NOTS
script `run_stan_freesigma_vec.R`. This prevents drift between the qmd
and the .R variants.

Key model features:

-   `log_beta ~ Normal(0, 4)` — weakly informative, \~5 orders of
    magnitude on β.
-   `sigma_obs_pl[p] ~ Half-Normal(0, 2)` (componentwise) — Gelman
    (2006) prior for variance components.
-   `sigma_obs_pl` is a **vector\[P\]**, one σ per plant. Cambio
    principal vs el modelo scalar de freesigma.
-   No `sigma_obs` data input, no `use_filter` flag.

```{r}
#| label: read-stan-model
stan_path <- file.path(proj_root, "stan", "renewal_A_fsv.stan")
stopifnot(file.exists(stan_path))

cat("Using Stan model:\n  ", stan_path, "\n", sep = "")

# Sanity preview: first 60 lines
readLines(stan_path) |> head(60) |> writeLines()
```

# 10 · Build the Stan data list {background-color="#2c3e50"}

*Paper §2.4–2.5 — assembling the inputs to the Bayesian renewal model.*

```{r}
#| label: build-stan-data
# y in P x T (Stan convention); existing y_batch is T x P.
y_PT <- t(y_batch)
stopifnot(identical(rownames(y_PT), Batch_focus),
          identical(rownames(W_geo), Batch_focus))

# Mask + dummy fill for NA cells (Stan rejects NA in `data`).
y_obs_mat <- matrix(as.integer(is.finite(y_PT) & y_PT > 0),
                    nrow = nrow(y_PT), ncol = ncol(y_PT))
y_filled  <- y_PT
y_filled[y_obs_mat == 0] <- 1.0

P <- length(Batch_focus)
T <- ncol(y_PT)

# NOTA: freesigma — NO se pasa sigma_obs ni use_filter al modelo.
# sigma_obs_pl es un parametro libre del Stan, no data.
build_stan_data <- function(rho) {
  list(
    P            = P,
    T            = T,
    G            = G,
    D            = D,
    g            = as.numeric(g_vec),
    d            = as.numeric(d_vec),
    W            = W_geo,
    rho          = rho,
    It_init      = as.numeric(It_init_vec),
    y            = y_filled,
    y_obs        = y_obs_mat,
    mean_gen_wk  = prior_mean_gen_days / 7,
    sd_gen_wk    = prior_sd_gen_days   / 7
  )
}

stan_data_demo <- build_stan_data(rho = rho_grid[1])
str(stan_data_demo, max.level = 1)
```

# 11 · Compile Stan model {background-color="#2c3e50"}

*Paper §2.5 — HMC settings (3 chains, 2000 warmup, 8000 sampling).*

```{r}
#| label: compile-stan
## stan_threads = TRUE matches the cached binary at stan/rsv_renewal_freesigma_vec,
## so cmdstanr reuses it instead of recompiling. Edit the .stan content and
## run `cmdstanr::rebuild_cmdstan()` once to refresh the PCH if needed.
mod <- cmdstan_model(stan_file = stan_path, compile = TRUE,
                     cpp_options = list(stan_threads = TRUE))

cat("Compiled binary at:", mod$exe_file(), "\n")
```

# 12 · Sample at ρ = 0 {background-color="#2c3e50"}

*Paper §3 — fits the Data variant with ρ = 0 (spatially independent baseline).*

## The sweep

We compile the freesigma_vec model once and call `$sample()` once per
rho. Each fit is persisted to
`outputs/stan_fits_fsv_4ch9k/rho_<NN>.rds`. Set
`RUN_SAMPLE <- FALSE` to skip the sampling.

```{r}
#| label: stan-sample-rho-grid
#| eval: true

RUN_SAMPLE <- TRUE

# IMPORTANTE: salida en stan_fits_fsv_4ch9k/ para NO sobreescribir
# los fits anteriores en stan_fits_fsv_v3/. Esta corrida usa la
# configuracion ess-boost: 4 chains x 9000 sampling = 36000 draws.
fits_dir <- file.path(proj_root, "outputs", "stan_fits_fsv_4ch9k")
dir.create(fits_dir, showWarnings = FALSE, recursive = TRUE)

# init: sigma_obs_pl ahora es VECTOR[P] (uno por planta).
init_fun <- function(chain_id) {
  list(
    log_beta     = rep(0, P),
    sigma_obs_pl = rep(1.0, P),          # VECTOR P, no escalar
    lam          = matrix(0.1, P, T),    # λ≈0.1 (R_t ≈ 1.12, casi neutral)
    It           = matrix(rep(It_init_vec, T), P, T)
  )
}
init_list <- lapply(seq_len(n_chains), init_fun)

fits <- vector("list", length(rho_grid))
names(fits) <- sprintf("rho_%.2f", rho_grid)

# ---- Worker function: ajusta un rho + guardado robusto -------------------
fit_one_rho_stan <- function(i, rho_grid, stan_file, build_stan_data, init_list,
                             n_chains, iter_warmup, iter_sampling,
                             adapt_delta, max_treedepth, fits_dir) {
  suppressPackageStartupMessages({
    library(cmdstanr); library(posterior)
  })
  mod <- cmdstan_model(stan_file,
                       compile = TRUE,
                       cpp_options = list(stan_threads = TRUE))
  this_rho <- rho_grid[i]
  pid <- Sys.getpid()
  cat(sprintf("[worker pid=%d] rho = %.2f  (%d/%d)  [freesigma]\n",
              pid, this_rho, i, length(rho_grid)))
  t0 <- Sys.time()
  sd_i <- build_stan_data(this_rho)
  fit  <- mod$sample(
    data              = sd_i,
    init              = init_list,
    chains            = n_chains,
    parallel_chains   = n_chains,
    threads_per_chain = 1,
    iter_warmup       = iter_warmup,
    iter_sampling     = iter_sampling,
    adapt_delta       = adapt_delta,
    max_treedepth     = max_treedepth,
    refresh           = 500,
    seed              = 9000 + as.integer(this_rho * 100)
  )

  # ---- ROBUST SAVE: extract draws + write to TMPDIR with xz + mv ----
  cat(sprintf("[worker pid=%d] rho = %.2f extracting draws...\n",
              pid, this_rho))

  # Incluye sigma_obs_pl en los draws para que el analisis lo reporte.
  draws_arr <- fit$draws(variables = c("log_beta", "beta",
                                        "sigma_obs_pl",
                                        "Rt", "It",
                                        "log_lik", "lp__"),
                          format = "draws_array")
  diag_sum  <- fit$diagnostic_summary()
  summary_tbl <- fit$summary(
    variables = c("log_beta", "beta", "sigma_obs_pl", "Rt", "It", "lp__"),
    rhat = posterior::rhat,
    ess_bulk = posterior::ess_bulk,
    ess_tail = posterior::ess_tail,
    mean, sd
  )

  save_object <- list(
    rho            = this_rho,
    draws          = draws_arr,
    diagnostic_summary = diag_sum,
    summary        = summary_tbl,
    meta           = list(
      model_variant  = "freesigma_vec_per_plant",
      n_chains       = n_chains,
      iter_warmup    = iter_warmup,
      iter_sampling  = iter_sampling,
      adapt_delta    = adapt_delta,
      max_treedepth  = max_treedepth,
      save_time      = Sys.time()
    )
  )

  rds_path <- file.path(fits_dir,
                        sprintf("rho_%02d.rds",
                                as.integer(round(100 * this_rho))))
  tmp_path <- file.path(Sys.getenv("TMPDIR", tempdir()),
                        sprintf("rho_%02d_%d.rds",
                                as.integer(round(100 * this_rho)), pid))
  # gzip en vez de xz: xz se corrompe si el proceso muere mid-write,
  # gzip es mas tolerante.
  saveRDS(save_object, tmp_path, compress = "gzip")

  # file.copy + remove (file.rename falla cross-filesystem en NOTS)
  ok <- file.copy(tmp_path, rds_path, overwrite = TRUE)
  if (!ok) stop(sprintf("file.copy failed: %s -> %s", tmp_path, rds_path))
  file.remove(tmp_path)
  if (!file.exists(rds_path) || file.size(rds_path) < 1024) {
    stop(sprintf("Final file missing or too small: %s", rds_path))
  }

  cat(sprintf("[worker pid=%d] done rho = %.2f in %.1f min -> %s\n",
              pid, this_rho,
              as.numeric(difftime(Sys.time(), t0, units = "mins")),
              basename(rds_path)))
  rds_path
}

if (RUN_SAMPLE) {
  stan_file_path <- mod$stan_file()

  slurm_cpus <- suppressWarnings(as.integer(Sys.getenv("SLURM_CPUS_PER_TASK", "")))
  detected   <- if (!is.na(slurm_cpus) && slurm_cpus > 0L) slurm_cpus
                else parallel::detectCores(logical = FALSE)
  # MODO DEBUG: forzar secuencial. Cambia n_workers > 1 para paralelo PSOCK.
  n_workers <- 1L
  # n_workers <- min(length(rho_grid),
  #                  max(1L, floor(detected / n_chains)))
  cat(sprintf("[rho-loop] %d rho values, %d worker(s) (SEQUENTIAL), %d chains each = %d cores\n",
              length(rho_grid), n_workers, n_chains, n_workers * n_chains))

  if (n_workers > 1L) {
    cl <- parallel::makeCluster(n_workers, type = "PSOCK", outfile = "")
    on.exit(try(parallel::stopCluster(cl), silent = TRUE), add = TRUE)
    parallel::clusterExport(cl, varlist = c(
      "P", "T", "G", "D",
      "g_vec", "d_vec", "W_geo",
      "It_init_vec", "y_filled", "y_obs_mat",
      "prior_mean_gen_days", "prior_sd_gen_days"
    ), envir = environment())
    paths <- parallel::parLapply(
      cl,
      seq_along(rho_grid),
      fit_one_rho_stan,
      rho_grid         = rho_grid,
      stan_file        = stan_file_path,
      build_stan_data  = build_stan_data,
      init_list        = init_list,
      n_chains         = n_chains,
      iter_warmup      = iter_warmup,
      iter_sampling    = iter_sampling,
      adapt_delta      = adapt_delta,
      max_treedepth    = max_treedepth,
      fits_dir         = fits_dir
    )
    parallel::stopCluster(cl)
  } else {
    paths <- lapply(seq_along(rho_grid), fit_one_rho_stan,
                    rho_grid         = rho_grid,
                    stan_file        = stan_file_path,
                    build_stan_data  = build_stan_data,
                    init_list        = init_list,
                    n_chains         = n_chains,
                    iter_warmup      = iter_warmup,
                    iter_sampling    = iter_sampling,
                    adapt_delta      = adapt_delta,
                    max_treedepth    = max_treedepth,
                    fits_dir         = fits_dir)
  }

  # Reload fits from disk into the named list.
  for (i in seq_along(paths)) fits[[i]] <- readRDS(paths[[i]])

  # Save the combined collection (mirror of NIMBLE's fits_all.rds).
  # Sufijo _4ch9k para distinguir de la corrida anterior fits_all_freesigma.rds.
  combined_path <- file.path(proj_root, "outputs",
                             "fits_all_freesigma_4ch9k.rds")
  saveRDS(fits, combined_path)
  cat("Saved", combined_path, "with", length(fits), "freesigma rho fits.\n")
}
```

# 13 · Diagnostics across rho {background-color="#2c3e50"}

## R-hat and ESS per rho (incluye `sigma_obs_pl`)

```{r}
#| label: diagnostics-table
#| eval: false
# DESACTIVADO: el save es formato lista (no R6). Para analisis post-fit,
# usa /Users/josepalacio/Documents/Claude/Projects/NOTS/analyze_stan_results_freesigma.qmd
# adaptado para leer outputs/stan_fits_fsv_4ch9k/ y la sigma por planta.
if (RUN_SAMPLE) {
  diag_df <- purrr::imap_dfr(fits, function(fit, key) {
    if (is.null(fit)) return(NULL)
    sm <- fit$summary
    # sigma_obs_pl ahora es vector[P] -> usamos median entre plantas.
    sigma_rows <- sm[grepl("^sigma_obs_pl", sm$variable), ]
    data.frame(
      rho               = as.numeric(sub("rho_", "", key)),
      rhat_max          = max(sm$rhat, na.rm = TRUE),
      ess_bulk_min      = min(sm$ess_bulk, na.rm = TRUE),
      ess_tail_min      = min(sm$ess_tail, na.rm = TRUE),
      n_divergent       = sum(fit$diagnostic_summary$num_divergent),
      sigma_obs_med_p   = median(sigma_rows$mean),
      sigma_obs_min_p   = min(sigma_rows$mean),
      sigma_obs_max_p   = max(sigma_rows$mean)
    )
  })

  knitr::kable(diag_df, digits = c(2, 3, 0, 0, 0, 3, 3, 3),
               caption = "Convergence + sigma_obs_pl (per-plant) summary, freesigma_vec.")
}
```

# 14 · Notes and known limitations {background-color="#2c3e50"}

-   **No MARSS Stage 1.** Esto es la diferencia clave vs el archivo
    canónico. La σ_obs ya no viene de un filtro estado-espacio; se
    estima directamente en Stan.
-   **σ_obs per-planta (vector\[P\])**: cambio principal vs el modelo
    scalar de freesigma. Cada planta tiene su propio σ_p para capturar
    la heterogeneidad real (noise floor empírico va de 0.47 a 0.98
    entre plantas). Esperado: mejora de coverage 95% empírico de \~50%
    (scalar) a \~85-95% (vec).
-   **Esperado para β**: los CIs de β_p van a CRECER vs el scalar
    freesigma (no encoger). La varianza que el modelo scalar absorbía
    en β se redistribuye hacia σ por planta. Eso es lo que queremos —
    los CIs ultra-estrechos del scalar freesigma son síntoma de
    sobreidentificación.
-   **Wall-clock cost.** With 3 chains × 2000 warmup × 8000 sampling y P
    = 15, cada rho toma \~1.5–4 h. El sweep de 1 rho = 1.5–4 h en
    secuencial. Esperado: ~9% más por chain que la corrida anterior
    (3 chains × 8000 sampling), pero ESS ~1.5× mayor.
-   **Output dir.** Los .rds van a `outputs/stan_fits_fsv_4ch9k/` y el
    combinado a `outputs/fits_all_freesigma_4ch9k.rds`, **separados de
    la corrida previa** en `outputs/stan_fits_fsv_v3/` y
    `outputs/fits_all_freesigma.rds`, que NO se sobreescriben. Para
    analizar este experimento, abre el analyze script de tu elección y
    apunta `FITS_DIR` a `file.path(PROJ_DIR, "outputs", "stan_fits_fsv_4ch9k")`.
