---
title: "Spatial RSV Renewal — Stan (MARSS rho-est AR(1), Beta(2,2) prior)"
subtitle: "Single-fit: rho estimated jointly, MARSS Stage 1 filtered input, AR(1) on lambda (Normal walk)"
author: "Jose Palacio"
date: today
date-format: "MMMM D, YYYY"
format:
  revealjs:
    theme: simple
    slide-number: true
    progress: true
    code-line-numbers: true
    highlight-style: github
    footer: "Spatial RSV Renewal · Stan · 15 plants · MARSS rho-est AR(1) · Beta(2,2)"
    transition: fade
    background-transition: fade
    incremental: false
    toc: false
    smaller: true
    scrollable: true
    fig-align: center
    fig-width: 9
    fig-height: 5
execute:
  echo: true
  eval: true
  warning: false
  message: false
  cache: false
knitr:
  opts_chunk:
    comment: "#>"
editor_options:
  chunk_output_type: console
editor: 
  markdown: 
    wrap: 72
---

```{=html}
<style>
:root { --font-scale: 0.75; --code-scale: 0.75; }
.reveal .slides         { font-size: calc(40px * var(--font-scale)); }
.reveal .slides h1      { font-size: calc(2.2em * var(--font-scale)); }
.reveal .slides h2      { font-size: calc(1.6em * var(--font-scale)); }
.reveal pre, .reveal pre code { font-size: calc(0.75em * var(--code-scale)); }
.reveal table           { font-size: calc(0.8em * var(--font-scale)); }
</style>
```

```{r}
#| label: root-dir
#| include: false
# NOTS bundle: proj_root = current working dir (rsv_mcmc_stan_geoW)
# proj_root: ruta absoluta a la carpeta donde viven data/, stan/, R/.
# Local Mac: usa absoluta para que funcione desde cualquier cwd.
# NOTS: cambiar a `normalizePath(".")` (cwd del SLURM submit dir).
proj_root <- if (Sys.info()["sysname"] == "Darwin") {
  "/Users/josepalacio/Documents/Claude/Projects/NOTS/rsv_mcmc_stan_geoW"
} else {
  normalizePath(".", mustWork = FALSE)
}
knitr::opts_knit$set(root.dir = proj_root)
cat("proj_root =", proj_root, "\n")
```

# 0 · What this file does {background-color="#2c3e50"}

## One file, no prerequisites

This `.qmd` is **self-contained**: every step from raw CSV/shapefile to
Stan posteriors lives here. No `stan_input_data.rds` to pre-cook.

The structure:

1.  **Data prep (lifted from `SPNIMBLE_GR_IN_presentation.qmd` sections
    0–8.5)** — reads the wastewater CSV + WWTP shapefile + census
    tracts, builds `y_batch`, `N_plantas`, `It_init_vec`, runs MARSS
    Stage 1 for `y_filt`/`sigma_filt`, and constructs the
    generation/shedding kernels `g_vec`/`d_vec`.
2.  **Buffer-capture analysis (lifted from `README.qmd`)** — for the 15
    WWTPs, builds the `d_pop_50` distance matrix (radius around each
    plant's centroid that captures 50 % of every other plant's
    population) via `region_distance_matrix()`, and prints the 15
    self-analysis plots from `buffer_capture_analysis()`.
3.  **Distance → W transformation (new)** — converts `d_pop_50` into a
    row-stochastic spatial weight matrix `W` via `W = K / rowSums(K)`
    with `K_ij = exp(-d_ij / sigma)` and `diag(K) = 0`.
4.  **Stan model + single fit** — compiles `renewal_A_marss_est.stan`
    once and runs **one fit** with ρ as a free parameter under a
    Beta(ALPHA_RHO, BETA_RHO) prior. No grid loop.

**The two knobs you will change most:**

``` r
Batch_focus <- paste0("WWTP", 1:15)
ALPHA_RHO   <- 2     # Beta(2, 2) prior shape parameter
BETA_RHO    <- 2     # Beta(2, 2) prior rate parameter
```

# 1 · Packages, paths, helpers {background-color="#2c3e50"}

## Packages

```{r}
#| label: packages
suppressPackageStartupMessages({
  library(tidyverse)
  library(lubridate)
  library(sf)
  library(MARSS)         # Stage 1 SSM (per-plant filter)
  library(cmdstanr)      # Stan interface
  library(posterior)     # draws_df, ess, rhat
  library(purrr)
})
set.seed(123)
```

## Settings — the only knobs

```{r}
#| label: settings
# ---- The two knobs that turn this run into "15 plants, rho sweep" ----------
Batch_focus <- paste0("WWTP", 1:15)
# Reverse order so the SLOW rhos start first when running PSOCK-parallel —
# MARSS rho-est AR(1): single fit, rho estimated with Beta(2, 2) prior.
ALPHA_RHO <- 2
BETA_RHO  <- 2
# Placeholder (downstream code expects rho_grid[1] in some debug paths)
rho_grid  <- c(NA_real_)

# W is built from the population-based distance matrix d_pop_50 only
# (centroid path removed for clarity).




# ---- Pipeline constants ----------------------------------------------------
casos_houston_semana <- 324
burnin               <- 15
cutoff_date_min      <- as.Date("2023-01-02")

prior_mean_gen_days  <- 7.5    # Vink 2014 RSV
prior_sd_gen_days    <- 2.1
mean_shed_days       <- 4.6    # Cevik 2023
sd_shed_days         <- 2.0

# ---- MCMC defaults ---------------------------------------------------------
n_chains      <- 3
iter_warmup   <- 2000
iter_sampling <- 8000
adapt_delta   <- 0.95
max_treedepth <- 15
```

## Helpers

```{r}
#| label: helpers
# to_gamma: Gamma shape/scale for a (mean, sd) given in DAYS, returned in WEEKS.
to_gamma <- function(mean_days, sd_days) {
  m <- mean_days / 7; s <- sd_days / 7
  list(shape = (m^2) / (s^2), scale = (s^2) / m)
}

# make_shedding_kernel: discretised Gamma shedding kernel including lag 0.
make_shedding_kernel <- function(mean_shed_days, sd_shed_days, K = 2) {
  par_d   <- to_gamma(mean_shed_days, sd_shed_days)
  edges_d <- c(-0.5, seq(0.5, K + 0.5, by = 1))
  F_d     <- pgamma(pmax(edges_d, 0),
                    shape = par_d$shape, scale = par_d$scale)
  w_d     <- diff(F_d)
  w_d     <- pmax(w_d, 0) / sum(pmax(w_d, 0))
  list(d = w_d, G = K, D = length(w_d))
}
```

# 2 · Shapefile + crosswalk {background-color="#2c3e50"}

*Paper §2.1 — Data and study area (population allocation via eq. 2).*

## Read the WWTP shapefile and build the WWTP1..WWTP15 crosswalk

The crosswalk is deterministic: unique `BatchName`s in the shapefile,
sorted alphabetically, are assigned `WWTP1, WWTP2, …` by `row_number()`.
This is what the NIMBLE pipeline does and what we need to match for
`y_batch` alignment.

```{r}
#| label: shapefile-crosswalk
stopifnot(file.exists(shp_batch))

wwtp_shp_batch <- sf::read_sf(shp_batch) |>
  sf::st_transform(32615) |>
  mutate(BatchName = toupper(trimws(BatchName)))

batch_key_shp <- wwtp_shp_batch |>
  sf::st_drop_geometry() |>
  distinct(BatchName) |>
  arrange(BatchName) |>
  mutate(Batch = paste0("WWTP", row_number()))

wwtp_shp_batch <- wwtp_shp_batch |>
  left_join(batch_key_shp, by = "BatchName")

wwtp_shp_focus <- wwtp_shp_batch |>
  filter(Batch %in% Batch_focus) |>
  arrange(match(Batch, Batch_focus))

stopifnot(nrow(wwtp_shp_focus) == length(Batch_focus))
n_plants <- length(Batch_focus)

cat("Crosswalk WWTPn -> BatchName:\n")
print(batch_key_shp |> filter(Batch %in% Batch_focus))
```

## Census-tract population allocation

```{r}
#| label: ct-allocation
ct_raw <- read_csv(csv_ct_pop, show_col_types = FALSE) |>
  mutate(BatchName = toupper(trimws(BatchName)))

ct_alloc <- ct_raw |>
  left_join(batch_key_shp, by = "BatchName") |>
  filter(!is.na(Batch)) |>
  mutate(
    pop_alloc = Percent_ce * POP24_5YR,
    moe_alloc = Percent_ce * MOE24_5YR
  )

# Geometry for census tracts (prefer real shapefile; fall back to synthetic).
if (file.exists(shp_ct_pop) &&
    grepl("\\.shp$", shp_ct_pop, ignore.case = TRUE)) {
  ct_sf <- sf::read_sf(shp_ct_pop) |>
    sf::st_transform(32615) |>
    mutate(BatchName = toupper(trimws(BatchName))) |>
    left_join(
      ct_alloc |> select(BatchName, Percent_ce, POP24_5YR, pop_alloc, Batch),
      by = c("BatchName", "Percent_ce", "POP24_5YR")
    ) |>
    filter(!is.na(Batch))
} else {
  message("Census tract shapefile not found — generating synthetic centroids.")
  ct_alloc_sf <- ct_alloc |>
    left_join(wwtp_shp_batch |> select(BatchName, geometry), by = "BatchName") |>
    sf::st_as_sf()
  set.seed(999)
  pts <- lapply(seq_len(nrow(ct_alloc_sf)), function(i) {
    s <- tryCatch(sf::st_sample(ct_alloc_sf[i, ], size = 1, type = "random"),
                  error = function(e) NULL)
    if (is.null(s) || length(s) == 0)
      s <- sf::st_point_on_surface(sf::st_geometry(ct_alloc_sf[i, ]))
    s
  })
  ct_sf <- ct_alloc |>
    mutate(geometry = do.call(c, pts)) |>
    sf::st_as_sf(crs = 32615)
}

ct_pop_by_batch <- ct_alloc |>
  group_by(Batch) |>
  summarise(N_acs = sum(pop_alloc, na.rm = TRUE),
            N_moe_acs = sqrt(sum(moe_alloc^2, na.rm = TRUE)),
            .groups = "drop")

cat("\nPopulation per batch:\n"); print(ct_pop_by_batch)
```

# 3 · Wastewater data {background-color="#2c3e50"}

*Paper §2.1 — viral load matrix (eq. 1).*

## Read CSV + apply the batch-code crosswalk

```{r}
#| label: wastewater-load
batch_crosswalk <- tibble(
  batch_code = c("B69", "BAS", "BIA", "BKW", "BNO", "BNE", "BNW",
                 "BSB", "BSO", "BSE", "BWE", "BWS", "BKB", "BSW", "BWB"),
  BatchName  = c("69TH STREET", "ALMEDA SIMS", "BATCH INTERCONTINENTAL",
                 "BATCH KINGWOOD", "BATCH NORTH", "BATCH NORTHEAST",
                 "BATCH NORTHWEST", "BATCH SIMS BAYOU", "BATCH SOUTH",
                 "BATCH SOUTHEAST", "BATCH WEST", "BATCH WEST SOUTH",
                 "KEEGANS BAYOU", "SOUTHWEST", "WILLOWBROOK")
)

ww_input <- read_csv(raw_csv_path, show_col_types = FALSE) |>
  mutate(WWTP  = toupper(trimws(WWTP)),
         tn    = toupper(trimws(tn)),
         BATCH = toupper(trimws(BATCH))) |>
  filter(WWTP != "SS", tn == "RSV") |>
  transmute(batch_code = BATCH, WWTP = WWTP, n_plants = Bnum,
            Date = as.Date(date), Pathogen = tn,
            copies_L = copies_L, Flow = flow_interp) |>
  arrange(batch_code, Date)

ww_input <- ww_input |>
  group_by(batch_code) |>
  mutate(median_flow = median(Flow, na.rm = TRUE)) |>
  ungroup() |>
  mutate(copies_day = (copies_L * median_flow) / 1e9) |>
  left_join(batch_crosswalk, by = "batch_code") |>
  left_join(batch_key_shp,   by = "BatchName")

batches_in_data <- ww_input |> filter(!is.na(Batch)) |>
  distinct(Batch) |> pull(Batch)
missing_focus <- setdiff(Batch_focus, batches_in_data)
if (length(missing_focus) > 0)
  stop("These Batch_focus plants have no wastewater data: ",
       paste(missing_focus, collapse = ", "),
       ". Either change Batch_focus or check the input CSV.")

ww <- ww_input |>
  filter(Batch %in% Batch_focus) |>
  arrange(Batch, Date)

cat("Plants present in CSV:", length(batches_in_data),
    " — focus uses:", length(Batch_focus), "\n")
```

# 4 · `N_plantas` and `It_init_vec` {background-color="#2c3e50"}

*Paper §2.1 (populations, eq. 2) and §2.4.4 (initial incidence anchor, eq. 17).*

```{r}
#| label: pop-and-It-init
N_plantas <- ct_pop_by_batch |>
  filter(Batch %in% Batch_focus) |>
  arrange(match(Batch, Batch_focus)) |>
  { \(d) setNames(d$N_acs, d$Batch) }()

N_houston <- sum(ct_pop_by_batch$N_acs, na.rm = TRUE)

# Allocate the weekly Houston-wide case count to each plant proportionally
# to its served population, then floor at 10. Floor reason: the Poisson →
# LogNormal moment-matched relaxation in Stan (mu_ln = log(lam) - 0.5/lam)
# becomes numerically delicate as lam → 0; a floor of 10 keeps every plant
# well inside the regime where the relaxation is accurate. The original
# pipeline (P=5) had natural It_init values of 20–80, so this floor only
# bites for the smaller plants among the 15 (Willowbrook, Batch North).
It_init_vec <- pmax(
  round(casos_houston_semana * N_plantas[Batch_focus] / N_houston),
  10
)
stopifnot(all(It_init_vec >= 10))

cat("N_plantas (1k):\n"); print(round(N_plantas / 1e3, 1))
cat("\nIt_init_vec:\n"); print(It_init_vec)
```

# 5 · `y_batch` — viral load matrix {background-color="#2c3e50"}

*Paper §2.1 — plant × week viral load matrix (eq. 1).*

```{r}
#| label: y-batch
ww_batch <- ww |>
  select(Batch, n_plants, Date, Pathogen, copies_day, flow_batch = Flow)

cutoff_date <- ww_batch |>
  summarise(d = dplyr::nth(sort(unique(Date)), burnin,
                           default = max(Date))) |>
  pull(d) |>
  max(cutoff_date_min)

ww_rsv_batch <- ww_batch |>
  filter(Date > cutoff_date) |>
  mutate(copies_day = ifelse(is.na(copies_day) | copies_day <= 0,
                              NA_real_, copies_day),
         Missing    = is.na(copies_day)) |>
  arrange(Batch, Date)

make_wide_matrix <- function(df, value_col, plants_order) {
  df |>
    mutate(.val = {{ value_col }}) |>
    select(Date, WWTP = Batch, .val) |>
    dplyr::group_by(Date, WWTP) |>
    dplyr::summarise(.val = if (all(is.na(.val))) NA_real_ else mean(.val, na.rm = TRUE),
                     .groups = "drop") |>
    pivot_wider(names_from = WWTP, values_from = .val, values_fill = NA_real_) |>
    arrange(Date) |>
    { \(w) {
      missing_cols <- setdiff(plants_order, names(w))
      for (m in missing_cols) w[[m]] <- NA_real_
      w |> select(all_of(c("Date", plants_order)))
    } }()
}

y_wide_batch <- make_wide_matrix(ww_rsv_batch, copies_day, Batch_focus)
y_batch      <- as.matrix(y_wide_batch[, -1, drop = FALSE])
rownames(y_batch) <- as.character(y_wide_batch$Date)

stopifnot(identical(colnames(y_batch), Batch_focus))

cat("y_batch:", nrow(y_batch), "weeks x", ncol(y_batch), "plants\n")
cat("Missing cells:", sum(is.na(y_batch)), "/",
    prod(dim(y_batch)), "\n")
```

# 6 · MARSS Stage 1 → `y_filt`, `sigma_filt` {background-color="#2c3e50"}

*Paper §2.6 — Two-stage strategy: state-space filter (eqs. 20–21) producing \`y_filt\` and \`sigma_filt\` as inputs to Stage 2.*

```{r}
#| label: marss-stage-1
ln10 <- log(10)
Y_log10_TP <- log10(y_batch)
Y_log10_TP[!is.finite(Y_log10_TP)] <- NA
P_focus <- ncol(Y_log10_TP)
T_focus <- nrow(Y_log10_TP)

ssm_spec_one_plant <- list(
  B  = matrix(c(2, 1, -1, 0), 2, 2, byrow = TRUE),
  U  = "zero", Q  = "diagonal and equal",
  V0 = "identity", x0 = "equal",
  R  = "diagonal and equal", A  = "zero",
  Z  = matrix(c(1, 0), 1, 2)
)

y_filt     <- matrix(NA_real_, P_focus, T_focus,
                     dimnames = list(Batch_focus, NULL))
sigma_filt <- matrix(NA_real_, P_focus, T_focus,
                     dimnames = list(Batch_focus, NULL))

for (p in seq_len(P_focus)) {
  yp_log10 <- Y_log10_TP[, p]
  if (sum(!is.na(yp_log10)) < 5) {
    warning(sprintf("Plant %s has <5 obs; skipping SSM.", Batch_focus[p]))
    y_filt[p, ] <- 1; sigma_filt[p, ] <- 1; next
  }
  fit_p <- MARSS::MARSS(matrix(yp_log10, nrow = 1),
                        model = ssm_spec_one_plant,
                        method = "BFGS", silent = TRUE,
                        control = list(maxit = 200))
  kfas_p <- MARSS::MARSSkfas(fit_p)
  xtt_p  <- kfas_p$xtt
  Vtt_p  <- kfas_p$Vtt
  Mp     <- coef(fit_p, type = "matrix")
  Zp     <- as.matrix(Mp$Z); Rp <- as.matrix(Mp$R)
  mu_log10_t <- as.numeric(Zp %*% xtt_p)
  V_obs_t    <- vapply(seq_len(T_focus),
                       function(tt) as.numeric(Zp %*% Vtt_p[, , tt] %*% t(Zp)),
                       numeric(1))
  V_obs_t <- pmax(V_obs_t, 0)
  mu_ln_t  <- ln10 * mu_log10_t
  var_ln_t <- (ln10^2) * V_obs_t
  y_filt[p, ]     <- exp(mu_ln_t + 0.5 * var_ln_t)
  sigma_filt[p, ] <- sqrt(var_ln_t)
}
# Cap sigma_filt to a sensible range. Very large values (which MARSS can
# return for plants with sparse data) make `mu_obs = log(psi+1e-6) - 0.5*s^2`
# blow up to -inf in Stan's likelihood, triggering "Location parameter is inf"
# rejections during warmup. A cap of 3 corresponds to a multiplicative factor
# of exp(3) ~ 20 — wider than any real RSV observation noise.
sigma_filt <- pmin(pmax(sigma_filt, 1e-6), 3.0)

cat("MARSS done. y_filt range: [",
    signif(min(y_filt, na.rm = TRUE), 3), ", ",
    signif(max(y_filt, na.rm = TRUE), 3), "]\n", sep = "")
```

```{r}
#| label: plot-filter-estimates-15
#| fig-width: 12
#| fig-height: 8

dates_vec <- as.Date(y_wide_batch$Date)

# One row per (plant, week): filtered mean + 95% lognormal CrI + observed dot
plot_df <- do.call(rbind, lapply(seq_along(Batch_focus), function(p) {
  psi   <- y_filt[p, ]
  sig   <- sigma_filt[p, ]
  mu_ln <- log(pmax(psi, 1e-300)) - 0.5 * sig^2

  data.frame(
    date     = dates_vec,
    plant    = Batch_focus[p],
    observed = as.numeric(y_batch[, p]),
    filtered = psi,
    lo       = exp(mu_ln + qnorm(0.025) * sig),
    hi       = exp(mu_ln + qnorm(0.975) * sig)
  )
}))
plot_df$plant <- factor(plot_df$plant, levels = Batch_focus)

p_filter <- ggplot(plot_df, aes(x = date)) +
  geom_ribbon(aes(ymin = lo, ymax = hi),
              fill = "steelblue", alpha = 0.25, na.rm = TRUE) +
  geom_line(aes(y = filtered), colour = "steelblue4",
            linewidth = 0.6, na.rm = TRUE) +
  geom_point(aes(y = observed), colour = "black",
             size = 0.7, na.rm = TRUE) +
  facet_wrap(~ plant, ncol = 3, nrow = 5, scales = "free_y") +
  labs(x = "Week", y = "Filter") +
  theme_bw(base_size = 11) +
  theme(strip.background = element_rect(fill = "grey92"),
        panel.grid.minor  = element_blank(),
        axis.text.x       = element_text(angle = 30, hjust = 1))

print(p_filter)

ggsave(file.path(proj_root, "filter_estimates_15.pdf"),
       p_filter, width = 12, height = 8, units = "in")
```

# 7 · Generation + shedding kernels {background-color="#2c3e50"}

*Paper §2.3 — generation and shedding kernels (eqs. 7–8).*

```{r}
#| label: kernels
K_shed <- 2     # shedding kernel support (in weeks beyond lag 0)
ks    <- make_shedding_kernel(mean_shed_days, sd_shed_days, K = K_shed)
d_vec <- as.numeric(ks$d)         # shedding (lag 0..K_shed)
G     <- ks$G                      # generation support length
D     <- ks$D

par_g <- to_gamma(prior_mean_gen_days, prior_sd_gen_days)
edges <- seq(1.5, G + 0.5, by = 1)
g_raw <- pgamma(edges, shape = par_g$shape, scale = par_g$scale) -
         pgamma(edges - 1, shape = par_g$shape, scale = par_g$scale)
g_raw <- pmax(g_raw, 0)
g_vec <- g_raw / sum(g_raw)        # generation kernel (sums to 1)

cat("g_vec (gen kernel, length", G, "):\n");  print(round(g_vec, 4))
cat("d_vec (shed kernel, length", D, "):\n"); print(round(d_vec, 4))
```

# 8 · Buffer-capture analysis on the 15 plants {background-color="#2c3e50"}

*Paper §2.2.1 — Population Extended Hausdorff Distance (eqs. 3–4).*

## Source helpers + read tracts (from `README.qmd`)

```{r}
#| label: source-buffer-capture
source(file.path(proj_root, "R", "core.R"))
source(file.path(proj_root, "R", "plot.R"))
source(file.path(proj_root, "R", "analysis.R"))

# Tracts layer for the buffer-capture analysis: read the project's tracts
# shapefile, dedupe by GEOID (it ships with one row per tract x batch), and
# project to the same CRS as wwtp_shp_focus (EPSG:32615 — UTM 15N, meters).
tracts_raw <- sf::st_read(tracts_shp_path, quiet = TRUE) |>
  sf::st_make_valid() |>
  dplyr::filter(sf::st_is_valid(geometry), !sf::st_is_empty(geometry))

tracts_valid <- tracts_raw |>
  dplyr::distinct(GEOID, .keep_all = TRUE) |>
  sf::st_transform(32615)

cat("tracts_valid:", nrow(tracts_valid), "unique tracts after dedupe.\n")
```

## Distance matrix: `d_pop_50`

```{r}
#| label: distance-matrices
# Both regions and tracts already in EPSG:32615 (meters).
d_pop_50 <- region_distance_matrix(
  regions         = wwtp_shp_focus,
  label_col       = "BatchName",
  method          = "population",
  tracts          = tracts_valid,
  pop_col         = "POP24_5YR",
  target_fraction = 0.5,
  people_per_dot  = 200,
  units           = "km"
)

cat("d_pop_50: ", paste(dim(d_pop_50), collapse = " x "), "\n", sep = "")
```

## Heatmap of the distance matrix

```{r}
#| label: distance-heatmaps
#| fig-width: 8
#| fig-height: 7
#| fig-show: "hide"
#| results: "hide"

# Guardamos a PDF directo para evitar el rendering inline de RStudio
# (que falla en notebooks sin guardar con 'QuartzBitmap_Output - unable to
# open file ...unsaved-notebooks/...'). Abre el PDF resultante para verlo.
dir.create("outputs/plots", showWarnings = FALSE, recursive = TRUE)
pdf("outputs/plots/distance_heatmap_pop50.pdf", width = 8, height = 7)
plot_distance_heatmap(d_pop_50, cluster = TRUE)
invisible(dev.off())
cat("→ outputs/plots/distance_heatmap_pop50.pdf\n")
```

## 15 self-analysis `buffer_capture_analysis()` plots

For each WWTP X we set `inner = outer = X` and ask: what's the radius
around X's centroid that captures 50 % of X's own population?

```{r}
#| label: buffer-capture-15
#| fig-width: 8
#| fig-height: 7
#| fig-show: "hide"
#| results: "hide"

# Same trick as distance-heatmaps / W-heatmaps: render to a PDF and hide
# the inline output. RStudio's inline renderer writes the chunk PNGs into
# ~/.local/share/rstudio/unsaved-notebooks/<id>/..., which on macOS fails
# with "QuartzBitmap_Output - unable to open file ..." whenever the .qmd
# isn't on disk yet, or whenever Quartz can't open that path. Going
# straight to PDF sidesteps the inline device entirely.
all_self_results <- purrr::set_names(unique(wwtp_shp_focus$BatchName)) |>
  purrr::map(function(bn) {
    one_batch <- dplyr::filter(wwtp_shp_focus, BatchName == bn)
    tryCatch(
      buffer_capture_analysis(
        inner           = one_batch,
        outer           = one_batch,
        tracts          = tracts_valid,
        proj_crs        = 32615,
        pop_col         = "POP24_5YR",
        labels          = list(inner = bn, outer = bn),
        people_per_dot  = 200,
        target_fraction = 0.5
      ),
      error = function(e) { warning(bn, ": ", conditionMessage(e)); NULL }
    )
  })

dir.create("outputs/plots", showWarnings = FALSE, recursive = TRUE)
pdf("outputs/plots/buffer_capture_self_15.pdf", width = 8, height = 7)
for (bn in names(all_self_results)) {
  res <- all_self_results[[bn]]
  if (is.null(res)) next
  print(res$plots$target_radius)
}
invisible(dev.off())
cat("→ outputs/plots/buffer_capture_self_15.pdf\n")
```

# 9 · Distance → spatial weight matrix W {background-color="#2c3e50"}

*Paper §2.2.2 — RBF kernel and row normalization (eqs. 5–6).*

## Why a distance is not a weight

The Stan model declares `matrix[P, P] W;` with the contract: `W` must be
**row-stochastic** (every row sums to 1, entries are non-negative) so
that
`I_mix[p, t] = (1 - rho) * It[p, t] + rho * sum_j W[p, j] * It[j, t]` is
a convex combination. A raw distance matrix `d` has `d[i, i] = 0`, large
positive off-diagonals, and asymmetric rows of different magnitudes —
none of which satisfies the contract.

The transformation we apply is the **Gaussian (RBF) kernel** with
bandwidth set by the **median heuristic on squared distances** — the
canonical default in the kernel-methods literature \[Gretton et al.,
2012, *JMLR* 13:723–773\]:

$$
K_{ij} \;=\; \exp\!\left(-\,\frac{d_{ij}^{\,2}}{2\sigma^{2}}\right),
\qquad
\sigma^{2} \;=\; \tfrac{1}{2}\,\text{median}\!\left(d_{ij}^{\,2}\,:\,i \neq j\right),
\qquad
K_{ii} \;\leftarrow\; 0, \qquad
W_{ij} \;=\; \frac{K_{ij}}{\sum_{k} K_{ik}}.
$$

Substituting the median into the kernel gives the equivalent form
$K_{ij} = \exp(-\,d_{ij}^{\,2}\,/\,\text{median}(d^{\,2}))$. The
bandwidth has the same units as $d_{ij}$ (km here). After the three
steps above:

-   `K` is non-negative, with `K_{ii} = 1` before we zero the diagonal.
-   Diagonal zeroed: no self-influence — a plant doesn't weight itself,
    only its neighbours via the `rho` mixing.
-   Each row divided by its row sum → `rowSums(W) == 1`.

**Why this bandwidth?** The median heuristic is scale-invariant, robust
to outlying plant pairs, and asymptotically consistent \[Garreau,
Jitkrittum & Kanagawa, 2017, arXiv:1707.07269\]. It is what authors mean
when they write "Gaussian kernel" in a methods section without further
qualification.

**Symmetry note.** Even when `d` is symmetric, row-normalisation breaks
symmetry — rows in dense regions get diluted differently from rows in
sparse regions. We accept this; the Stan model only requires
row-stochasticity, not double stochasticity.

```{r}
#| label: d-to-W
# Helper: row-stochastic W from a distance matrix `d` via Gaussian-like kernel.
# - sigma defaults to median(off-diagonal d). Pass an explicit value to override.
# - rownames/colnames of `d` are preserved in W.
make_W_from_distance <- function(d, sigma = NULL) {
  stopifnot(nrow(d) == ncol(d))

  # GAUSSIAN (RBF) KERNEL with bandwidth from the *median heuristic on
  # squared distances*:
  #
  #   K_{ij} = exp(-d_{ij}^2 / (2 * sigma^2))
  #   sigma^2 = median(d_{ij}^2 : i != j) / 2
  #   =>   K_{ij} = exp(-d_{ij}^2 / median(d_{ij}^2))
  #
  # Reference: Gretton, A., Borgwardt, K. M., Rasch, M. J., Schölkopf, B.,
  # & Smola, A. (2012). "A kernel two-sample test." Journal of Machine
  # Learning Research, 13:723-773. (Section 8.) This is the canonical
  # default in kernel methods and is what most authors mean when they
  # write "Gaussian kernel" without further qualification. Asymptotic
  # consistency: Garreau, Jitkrittum & Kanagawa (2017), arXiv:1707.07269.
  #
  # FIX (preserved from earlier): region_distance_matrix() returns a matrix
  # that still carries the "units" class from sf::st_distance(). exp() and
  # `/` on a units-classed matrix call `units(e1)$numerator` and crash with
  # "$ operator is invalid for atomic vectors". Strip the class baggage by
  # rebuilding the matrix as plain numeric.
  d_dim   <- dim(d)
  d_names <- dimnames(d)
  d <- structure(as.numeric(unclass(d)),
                 dim = d_dim, dimnames = d_names)

  off <- d
  diag(off) <- NA
  if (is.null(sigma)) {
    # Median heuristic on SQUARED distances (Gretton et al. 2012).
    sigma <- sqrt(median(off^2, na.rm = TRUE) / 2)
  }

  K <- exp(-d^2 / (2 * sigma^2))
  diag(K) <- 0
  rs <- rowSums(K)
  if (any(rs == 0))
    stop("Row sum of K is zero — pick a larger sigma or check the input matrix.")
  W <- K / rs
  attr(W, "sigma_km") <- sigma     # bandwidth used (km), = sqrt(median(d^2)/2)
  attr(W, "kernel")   <- "gaussian"
  W
}

# Apply to the population-based distance matrix. With no `sigma` argument,
# the function uses the Gaussian-kernel median heuristic on squared
# distances (Gretton et al. 2012). The exact bandwidth chosen is reported
# below and is a deterministic function of d_pop_50.
W_pop50 <- make_W_from_distance(d_pop_50)

cat(sprintf("kernel: %s, sigma (km): %.2f\n",
            attr(W_pop50, "kernel"),
            attr(W_pop50, "sigma_km")))

# Sanity: every row sums to 1, every diagonal is 0.
stopifnot(all(abs(rowSums(W_pop50) - 1) < 1e-10),
          all(diag(W_pop50) == 0))
```

## Heatmap of the resulting `W`

```{r}
#| label: W-heatmaps
#| fig-width: 8
#| fig-height: 7
#| fig-show: "hide"
#| results: "hide"

attr(W_pop50, "units")  <- "weight"
attr(W_pop50, "method") <- "pop50"
# digits = 3 so the tiny weights (typical: 0.02-0.20) actually render — with
# the default digits=0 every cell shows as "0" even though the colours differ.
# Guardamos a PDF para evitar el QuartzBitmap_Output error de RStudio inline.
dir.create("outputs/plots", showWarnings = FALSE, recursive = TRUE)
pdf("outputs/plots/W_heatmap_pop50.pdf", width = 8, height = 7)
plot_distance_heatmap(W_pop50, cluster = TRUE, digits = 3,
                      title = "W from population-50% distance (row-stochastic)")
invisible(dev.off())
cat("→ outputs/plots/W_heatmap_pop50.pdf\n")
```

## Rename BatchName → WWTPn and align with `Batch_focus`

```{r}
#| label: rename-W-WWTP
# The buffer-capture matrix uses BatchName ("69TH STREET", ...). The Stan
# data list expects rows/cols ordered identically to Batch_focus ("WWTP1"..)
# and identically to colnames(y_batch). Translate using batch_key_shp.
bn_to_wwtp <- setNames(batch_key_shp$Batch, batch_key_shp$BatchName)
stopifnot(all(rownames(W_pop50) %in% names(bn_to_wwtp)))

rename_to_wwtp <- function(W) {
  rownames(W) <- bn_to_wwtp[rownames(W)]
  colnames(W) <- bn_to_wwtp[colnames(W)]
  W[Batch_focus, Batch_focus, drop = FALSE]
}
W_pop50 <- rename_to_wwtp(W_pop50)

# The W that goes into Stan.
W_geo <- W_pop50

stopifnot(identical(rownames(W_geo), Batch_focus),
          identical(colnames(W_geo), Batch_focus),
          identical(rownames(W_geo), colnames(y_batch)))

cat(sprintf("W_geo: %d x %d (source = pop50)\n",
            nrow(W_geo), ncol(W_geo)))
```

# 10 · Stan model {background-color="#2c3e50"}

*Paper §2.4 (Model specification) and §2.5 (Priors, including Beta(2,2) on ρ).*

## Read the `.stan` file from disk

The source of truth for the MARSS variant with $\rho$ estimated is
`stan/renewal_A_marss_est.stan`. We read it from disk rather than
inlining it here to prevent drift between the QMD and the SLURM
runner (`run_A_marss_est.R`), which uses the same file.

Model features (see paper §2 for full specification):

- `log_beta ~ Normal(0, 4²)`, `sigma_obs_pl ~ Half-Normal(0, 2²)`
  componentwise, `rho ~ Beta(alpha_rho, beta_rho)` (default Beta(2, 2)).
- `use_filter = 1` in the data block routes `sigma_obs[p, t]` (from
  MARSS Stage 1) into the likelihood; `sigma_obs_pl` is present but
  inactive when `use_filter = 1`.

```{r}
#| label: read-stan-model
stan_path <- file.path(proj_root, "stan", "renewal_A_marss_est.stan")
stopifnot(file.exists(stan_path))
cat("Using Stan model:\n  ", stan_path, "\n", sep = "")

# Sanity preview: first 60 lines
readLines(stan_path) |> head(60) |> writeLines()
```

# 11 · Build the Stan data list {background-color="#2c3e50"}

*Paper §2.4–2.5 — assembling the inputs to the Bayesian renewal model.*

```{r}
#| label: build-stan-data
# y in P x T (Stan convention); existing y_batch is T x P.
y_PT <- t(y_batch)
stopifnot(identical(rownames(y_PT), Batch_focus),
          identical(rownames(W_geo), Batch_focus))

# ============================================================================
# CAMBIO ENSOR-2025: usar y_filt (MARSS-imputado) como input a Stan en lugar
# de y_raw con NAs reemplazados por dummy 1.0.
#
# Filosofia: MARSS Stage 1 es el "measurement model" que separa la true RNA
# load del ruido de PCR. Stan + renewal modela la dinamica epidemiologica
# sobre esa cantidad latente. Pasar y_filt (no y_raw) a Stan es coherente con
# esta arquitectura two-stage.
#
# Implicaciones:
#   - y_filt nunca es NA -> y_obs = 1 en TODAS las 1875 celdas (15 x 125)
#   - Likelihood activa en TODO -> mas info -> mejor convergencia
#   - ELPD absoluto CAMBIA vs version anterior (mas celdas en likelihood)
#   - El ranking entre rho deberia conservarse
# ============================================================================
y_filled  <- y_filt                                         # MARSS-imputado
y_obs_mat <- matrix(1L, nrow = nrow(y_PT), ncol = ncol(y_PT))  # likelihood en TODO

# Diagnostic: cuantas celdas tendrian y_raw real
y_obs_raw <- matrix(as.integer(is.finite(y_PT) & y_PT > 0),
                    nrow = nrow(y_PT), ncol = ncol(y_PT))
cat(sprintf("Likelihood activa en %d / %d cells (Ensor-style: y_filt como obs)\n",
            sum(y_obs_mat), prod(dim(y_obs_mat))))
cat(sprintf("  Detalle: %d cells con y_raw real, %d cells MARSS-imputado\n",
            sum(y_obs_raw), sum(y_obs_mat) - sum(y_obs_raw)))

P <- length(Batch_focus)
T <- ncol(y_PT)

mode_filt     <- TRUE
sigma_obs_mat <- if (mode_filt) sigma_filt else matrix(0.5, P, T)

build_stan_data <- function(...) {
  list(
    P            = P,
    T            = T,
    G            = G,
    D            = D,
    g            = as.numeric(g_vec),
    d            = as.numeric(d_vec),
    W            = W_geo,
    # rho REMOVED — now estimated as parameter
    It_init      = as.numeric(It_init_vec),
    y            = y_filled,
    y_obs        = y_obs_mat,
    mean_gen_wk  = prior_mean_gen_days / 7,
    sd_gen_wk    = prior_sd_gen_days   / 7,
    use_filter   = as.integer(mode_filt),
    sigma_obs    = sigma_obs_mat,
    # NEW: hyperparameters for Beta(alpha_rho, beta_rho) prior on rho
    alpha_rho    = ALPHA_RHO,
    beta_rho     = BETA_RHO
  )
}

stan_data_demo <- build_stan_data()
str(stan_data_demo, max.level = 1)
cat(sprintf("\nPrior on rho: Beta(%g, %g)  -> mean = %.3f, sd = %.3f\n",
            ALPHA_RHO, BETA_RHO,
            ALPHA_RHO / (ALPHA_RHO + BETA_RHO),
            sqrt(ALPHA_RHO * BETA_RHO /
                 ((ALPHA_RHO + BETA_RHO)^2 * (ALPHA_RHO + BETA_RHO + 1)))))
```

# 12 · Compile Stan model {background-color="#2c3e50"}

*Paper §2.5 — HMC settings (3 chains, 2000 warmup, 8000 sampling).*

```{r}
#| label: compile-stan
## We keep `stan_threads = TRUE` in cpp_options for one reason: it matches the
## options used to build the binary cached at `stan/rsv_renewal`, so cmdstanr
## reuses that binary instead of invoking `clang` to recompile. On a Mac whose
## SDK has moved from 14.0 to 14.4 (Xcode update) the CmdStan PCH was built
## for the old SDK and a fresh compile errors out with:
##   "PCH file was compiled for the target 'arm64-apple-macosx14.0.0' but the
##    current translation unit is being compiled for target 'arm64-apple-macosx14.4.0'"
## Cache hit avoids that path entirely. If you ever edit the .stan content, run
## `cmdstanr::rebuild_cmdstan()` once to refresh the PCH and the recompile will
## work again.
mod <- cmdstan_model(stan_file = stan_path, compile = TRUE,
                     cpp_options = list(stan_threads = TRUE))

# Show the first 40 lines of the compiled .stan as a sanity check.
# (mod$print() in cmdstanr prints by side-effect and returns the R6 object,
# so we read the file directly instead.)
readLines(stan_path) |> head(40) |> writeLines()
```

# 13 · Single fit — rho estimated jointly {background-color="#2c3e50"}

*Paper §3 — fits the Filter variant with ρ estimated jointly under Beta(2,2) prior.*

## The single-fit run

Unlike the sweep version, this experiment fits the model **once** with
$\rho$ as a parameter (Beta(`ALPHA_RHO`, `BETA_RHO`) prior). The fit is
persisted to `outputs/stan_fits_marss_rhoest/rho_estimated.rds`. Set
`RUN_SAMPLE <- FALSE` to skip sampling (useful while you debug data prep).

```{r}
#| label: stan-sample-rhoest
#| eval: true

RUN_SAMPLE <- TRUE

fits_dir <- file.path(proj_root, "outputs", "stan_fits_marss_rhoest")
dir.create(fits_dir, showWarnings = FALSE, recursive = TRUE)

# init: sigma_obs_pl VECTOR[P] (no usado por likelihood con use_filter=1
# pero igual existe como parametro), rho centrado en 0.5 (prior mean).
init_fun <- function(chain_id) {
  list(
    log_beta     = rep(0, P),
    sigma_obs_pl = rep(0.5, P),
    lam          = matrix(0.1, P, T),
    It           = matrix(rep(It_init_vec, T), P, T),
    rho          = 0.5   # arrancar en el centro del prior
  )
}
init_list <- lapply(seq_len(n_chains), init_fun)

if (RUN_SAMPLE) {
  cat(sprintf("Sampling ONCE — MARSS rho-est AR(1) with Beta(%g, %g) prior on rho\n",
              ALPHA_RHO, BETA_RHO))
  t0 <- Sys.time()

  sd_one <- build_stan_data()
  fit <- mod$sample(
    data              = sd_one,
    init              = init_list,
    chains            = n_chains,
    parallel_chains   = n_chains,
    threads_per_chain = 1,
    iter_warmup       = iter_warmup,
    iter_sampling     = iter_sampling,
    adapt_delta       = adapt_delta,
    max_treedepth     = max_treedepth,
    refresh           = 500,
    seed              = 1729
  )

  cat(sprintf("\nSampling done in %.1f min\n",
              as.numeric(difftime(Sys.time(), t0, units = "mins"))))

  cat("Extracting draws (includes rho) ...\n")
  draws_arr <- fit$draws(
    variables = c("log_beta", "beta", "sigma_obs_pl", "Rt", "It",
                  "rho", "log_lik", "lp__"),
    format = "draws_array")
  diag_sum  <- fit$diagnostic_summary()
  summary_tbl <- fit$summary(
    variables = c("log_beta", "beta", "sigma_obs_pl", "Rt", "It",
                  "rho", "lp__"),
    rhat = posterior::rhat,
    ess_bulk = posterior::ess_bulk,
    ess_tail = posterior::ess_tail,
    mean, sd,
    q5  = ~ quantile(.x, 0.05),
    q50 = ~ quantile(.x, 0.50),
    q95 = ~ quantile(.x, 0.95)
  )

  rho_post <- summary_tbl[summary_tbl$variable == "rho", ]
  cat("\n=== POSTERIOR OF RHO ===\n")
  print(rho_post)

  save_object <- list(
    rho_prior             = list(family = "beta",
                                 alpha  = ALPHA_RHO,
                                 beta   = BETA_RHO),
    rho_posterior_summary = rho_post,
    model_variant         = "MARSS_hybrid_rhoest",
    draws                 = draws_arr,
    diagnostic_summary    = diag_sum,
    summary               = summary_tbl,
    y_raw                 = y_PT,
    y_filt                = y_filt,
    sigma_filt            = sigma_filt,
    y_obs_mask            = y_obs_raw,
    d_vec                 = d_vec,
    meta = list(
      n_chains        = n_chains,
      iter_warmup     = iter_warmup,
      iter_sampling   = iter_sampling,
      adapt_delta     = adapt_delta,
      max_treedepth   = max_treedepth,
      log_beta_prior  = "N(0, 4)",
      sigma_obs_prior = "Half-Normal(0, 2) per plant  (no usado: use_filter=1)",
      sigma_obs_data  = "MARSS Stage 1 per-(p,t) variance",
      rho_prior       = sprintf("Beta(%g, %g)", ALPHA_RHO, BETA_RHO),
      ar_order        = "AR(1) on lambda (Normal walk)",
      save_time       = Sys.time()
    )
  )

  rds_path <- file.path(fits_dir, "rho_estimated.rds")
  tmp_path <- file.path(Sys.getenv("TMPDIR", tempdir()),
                        sprintf("rho_estimated_%d.rds", Sys.getpid()))
  saveRDS(save_object, tmp_path, compress = "gzip")
  ok <- file.copy(tmp_path, rds_path, overwrite = TRUE)
  if (!ok) stop("file.copy failed")
  file.remove(tmp_path)
  cat(sprintf("Saved fit -> %s  (%.1f MB)\n",
              rds_path, file.size(rds_path) / 1e6))

  fits <- list(rho_estimated = save_object)
} else {
  rds_path <- file.path(fits_dir, "rho_estimated.rds")
  if (file.exists(rds_path)) {
    fits <- list(rho_estimated = readRDS(rds_path))
    cat("Loaded existing fit:", rds_path, "\n")
  } else {
    stop("No existing fit and RUN_SAMPLE = FALSE.")
  }
}
```

# 14 · Convergence + rho posterior diagnostics {background-color="#2c3e50"}

*Paper §3.1 (Convergence diagnostics) and §3.2 (Posterior of ρ).*

## R-hat, ESS, divergences + posterior summary of $\rho$

```{r}
#| label: diagnostics-table
#| eval: false
# DESACTIVADO: el save ahora es formato lista (no R6) para compatibilidad
# con NOTS. Las secciones downstream que usaban fit$summary(), $draws(),
# $diagnostic_summary() ya no funcionan. Para el análisis post-fit,
# usa /Users/josepalacio/Documents/Claude/Projects/NOTS/analyze_stan_results.qmd
# que ya está adaptado para detectar y leer el formato nuevo.
if (RUN_SAMPLE) {
  diag_df <- purrr::imap_dfr(fits, function(fit, key) {
    if (is.null(fit)) return(NULL)
    sm <- fit$summary(
      variables = c("log_beta", "Rt", "It", "beta", "lp__"),
      .num_args = list(notation = "dec", digits = 4)
    )
    data.frame(
      rho            = as.numeric(sub("rho_", "", key)),
      rhat_max       = max(sm$rhat, na.rm = TRUE),
      ess_bulk_min   = min(sm$ess_bulk, na.rm = TRUE),
      ess_tail_min   = min(sm$ess_tail, na.rm = TRUE),
      n_divergent    = sum(fit$diagnostic_summary()$num_divergent),
      lp_mean        = mean(fit$draws("lp__", format = "draws_matrix"))
    )
  })

  knitr::kable(diag_df, digits = c(2, 3, 0, 0, 0, 1),
               caption = "Convergence diagnostics by rho.")
}
```

## Posterior `Rt` overlays per plant, coloured by rho

```{r}
#| label: rt-overlay-rho
#| eval: false
#| fig-width: 10
#| fig-height: 7
# DESACTIVADO — usar analyze_stan_results.qmd para los plots de Rt.
if (RUN_SAMPLE) {
  extract_Rt_summary <- function(fit, rho_val, plant_names) {
    sm <- fit$summary("Rt", mean = mean,
                      q025 = ~quantile(.x, 0.025, na.rm = TRUE),
                      q975 = ~quantile(.x, 0.975, na.rm = TRUE))
    idx <- stringr::str_match(sm$variable, "Rt\\[(\\d+),(\\d+)\\]")
    sm$p <- as.integer(idx[, 2]); sm$t <- as.integer(idx[, 3])
    sm$plant <- plant_names[sm$p]
    sm$rho   <- rho_val
    sm
  }
  rt_long <- purrr::imap_dfr(fits, function(fit, key) {
    if (is.null(fit)) return(NULL)
    extract_Rt_summary(fit,
                       rho_val = as.numeric(sub("rho_", "", key)),
                       plant_names = Batch_focus)
  })
  ggplot(rt_long, aes(x = t, y = mean,
                      group = rho, colour = factor(rho))) +
    geom_line(alpha = 0.8) +
    facet_wrap(~ plant, scales = "free_y") +
    labs(x = "week", y = "Rt posterior mean",
         colour = "rho",
         title = "Rt(t) by plant, one line per rho in the sweep") +
    theme_minimal(base_size = 10) +
    theme(legend.position = "bottom")
}
```

## LOO comparison across rho

```{r}
#| label: loo-compare
#| eval: false
# DESACTIVADO — usar analyze_stan_results.qmd para LOO comparison.
if (RUN_SAMPLE) {
  loo_list <- purrr::imap(fits, function(fit, key) {
    if (is.null(fit)) return(NULL)
    ll <- fit$draws("log_lik", format = "draws_matrix")
    # collapse the matrix of log_lik entries to a (draws x N_obs) matrix
    ll_mat <- as.matrix(ll)
    loo::loo(ll_mat, r_eff = NA)
  })
  loo_list <- Filter(Negate(is.null), loo_list)
  if (length(loo_list) >= 2) {
    print(loo::loo_compare(loo_list))
  } else {
    cat("Need at least 2 fits with log_lik for loo_compare.\n")
  }
}
```

# 15 · Notes and known limitations {background-color="#2c3e50"}

-   **`mode_filt = TRUE`** is hard-coded for the sweep. To compare
    against the Mode A (single sigma per plant), set
    `mode_filt <- FALSE` in the "Build the Stan data list" chunk and
    re-render section 13.
-   **Wall-clock cost.** With 3 chains × 2000 warmup × 8000 sampling and
    P = 15, each rho takes \~15–30 min. The full sweep = 2.5–5 h. For a
    smoke run, drop `iter_warmup` / `iter_sampling` to 250.
-   **`W` source.** Built from `d_pop_50` only (population-based,
    asymmetric). If later you want the centroid-based version back, add
    a parallel call `W_centroid <- make_W_from_distance(d_centroid)` and
    rename rows the same way; the rest of the pipeline downstream is
    W-source-agnostic.
-   **Population-tract dedupe.** The buffer-capture analysis dedupes the
    tracts layer by `GEOID` before sampling, so tracts that span
    multiple batches in the original shapefile aren't double-counted in
    the dot mesh.
