---
title: "README"
format: html
---

# Buffer-capture analysis

How well does a circular buffer around one polygon approximate another, nesting one --- in terms of captured population? One function, one result list.

## Layout

```         
proj/
├── R/
│   ├── core.R                  functions for computing distances
│   ├── plot.R                  plot helpers + default colors / labels
│   ├── analysis.R              orchestrator + save helpers
│   ├── generate_test_data.R    states + counties test for example (fake population data)
│   └── source_all.R            sources the four files above
├── outputs/                    serialised result + bisection log
├── figures/                    PNG plots
├── analysis.qmd                narrative report
└── review_plan.md              what to check, why, and how
```

## Syntax 

You need two `sf` polygons --- the smaller one nesting inside the larger --- and a tract layer with a population column. The function takes them directly; nothing has to be renamed.

The following functions are defined in `R/core.r`. See below for example usage.

``` r
result <- buffer_capture_analysis(
  inner           = my_inner_batch,    # sf, single row, smaller
  outer           = my_outer_batch,    # sf, single row, contains inner
  tracts          = my_tracts,         # sf with a population column
  pop_col         = "Pop",
  people_per_dot  = 10,
  target_fraction = 0.95,
  tolerance       = list(frac = 0.005, radius_m = 25),
  proj_crs        = 32615,             # any projected CRS in meters
  colors          = list(              # any subset; defaults fill in
    inner_fill   = "#f5d4cf",
    outer_stroke = "#1f6fa8",
    dot_inside   = "#7a4a00"
  ),
  labels          = list(inner = "69th Street", outer = "Sims Bayou")
)

# Inspect anything:
result$centroid_guess$proportion
result$target$radius
result$mesh$recommended$k_safe
result$plots$target_radius

# Optionally save to disk:
save_buffer_analysis(result, fig_dir = "figures", out_dir = "outputs")

# Render the report:
quarto::quarto_render("analysis.qmd")
```

## Arguments

| Argument | Default | What it does |
|------------------------|------------------------|------------------------|
| `inner` | -- | Single-row sf polygon. Buffer is centered on its centroid. |
| `outer` | -- | Single-row sf polygon. Population is sampled inside it. Assumed to contain `inner`. |
| `tracts` | -- | sf polygon layer with a population column. Anything outside `outer` is filtered out. |
| `pop_col` | `"Pop"` | Name of the population column in `tracts`. |
| `people_per_dot` | `10` | Approx. people per grid point. Smaller -\> denser mesh, slower, more accurate. |
| `target_fraction` | `0.95` | Captured fraction the optimised buffer must hit. |
| `tolerance` | `list(frac=0.005, radius_m=25)` | Bisection stopping criteria. Pass partial; defaults fill in. |
| `proj_crs` | `NULL` | EPSG code or CRS object. If non-NULL, all inputs are reprojected before any geometry math. Use a projected CRS in meters. |
| `rule` | `"centroid_in"` | Tract-in-batch rule. `"intersects"` for an inclusive variant. |
| `colors` | `NULL` | Partial override of `default_colors()`. See `R/plot.R` for the full key set. |
| `labels` | `NULL` | Partial override of `default_labels()` (`inner`, `outer`). Used in plot titles. |
| `seed` | `20250507` | Reproducibility seed. |

## What you get back

A list with these slots:

```         
$inputs           the resolved inputs (in projected CRS)
$dots             sf points; one row per population dot
$centroid_guess   $center, $cent_outer, $radius, $proportion,
                  $buffer, $intersection, $captured
$target           $target_fraction, $radius, $proportion,
                  $iterations, $iter_log,
                  $buffer, $intersection, $captured, $center
$mesh             $geometry, $current, $tolerances, $recommended,
                  $meets_target
$plots            $overview, $dots, $centroid_guess, $target_radius
                  (all ggplot objects)
```

## Try it with fake population data and state boundaries

`generate_test_data()` builds a state-and-county fixture with synthetic populations:

```{r}
source("R/source_all.R")

td <- generate_test_data("LA", "TX")        # any two state codes

# Real US states don't truly nest, so for a clean demo of the nesting
# assumption, wrap `inner` in a buffer to manufacture an `outer`:
#fake_outer <- make_nesting_outer(td$inner, buffer_km = 200)

result <- buffer_capture_analysis(
  inner    = td$inner,
  outer    = td$outer,
  tracts   = td$tracts,
  proj_crs = 5070,                          # CONUS Albers
  labels   = list(inner = "Louisiana", outer = "LA + 200km buffer"),
  people_per_dot = 500000
)

result$plots
```

Or, to exercise the geometry on overlapping (not strictly nesting) shapes, pass the two states straight through. The math still runs; you just lose the "buffer is bounded by outer's perimeter" intuition.

## Picking proj_crs

Buffer distances need a CRS in meters. Common choices:

-   `5070` (NAD83 / Conus Albers Equal Area) -- safe default for any pair of contiguous-US polygons.
-   `32615` (UTM 15N) -- Houston, eastern Texas, Louisiana.
-   `32610`-`32619` -- other UTM zones for North America.
-   Whatever your data ships with, if it's already projected.

A geographic CRS (lat/lon) is rejected with a warning; the function will not produce sensible distances.

## Try with WWTP batch data

```{r}
## you can download the pop density data from rebecca's email-- put that folder in the same one as this file.
## note: you will have to update the paths in your version of Read in shape files.R
source("23. Population Density WWTP/Read in shape files.R") 

## add code to intersect overlapped tracts and split for true nesting

result <- buffer_capture_analysis(
  inner    = dplyr::filter(batches_valid, BatchName == "BATCH SIMS BAYOU"),
  outer    = dplyr::filter(batches_valid, BatchName == "69TH STREET"),
  tracts   = tracts_valid,
  proj_crs = 5070,# CONUS Albers
  pop_col  = "POP24_5YR",
  labels   = list(inner = "Batch Sims Bayou", outer = "69th St"),
  people_per_dot = 200, 
  target_fraction = .5 ## median
)

result$plots
```

## Compute distance matrix

`region_distance_matrix()` refuses geographic CRS, so project once up front and reuse for both calls below.

### Centroid

Centroid-to-centroid distance between every pair of batches:

```{r}
batches_proj <- sf::st_transform(batches_valid, 5070)
tracts_proj  <- sf::st_transform(tracts_valid,  5070)

d_centroid <- region_distance_matrix(
  regions   = batches_proj,
  label_col = "BatchName",
  method    = "centroid",
  units     = "km"
)
plot_distance_heatmap(d_centroid, cluster = TRUE)
```

### Population %

Population-based: entry `[i, j]` is the radius around batch i's centroid that captures half (`target_fraction = 0.5`) of batch j's population, using the same tract layer, column, and `people_per_dot` as the buffer analysis above.

```{r}
d_pop_50 <- region_distance_matrix(
  regions         = batches_proj,
  label_col       = "BatchName",
  method          = "population",
  tracts          = tracts_proj,
  pop_col         = "POP24_5YR",
  target_fraction = 0.5,
  people_per_dot  = 200,
  units           = "km"
)
plot_distance_heatmap(d_pop_50, cluster = TRUE)

d_pop_25 <- region_distance_matrix(
  regions         = batches_proj,
  label_col       = "BatchName",
  method          = "population",
  tracts          = tracts_proj,
  pop_col         = "POP24_5YR",
  target_fraction = 0.25,
  people_per_dot  = 200,
  units           = "km"
)
plot_distance_heatmap(d_pop_25, cluster = TRUE)

d_pop_75 <- region_distance_matrix(
  regions         = batches_proj,
  label_col       = "BatchName",
  method          = "population",
  tracts          = tracts_proj,
  pop_col         = "POP24_5YR",
  target_fraction = 0.75,
  people_per_dot  = 200,
  units           = "km"
)
plot_distance_heatmap(d_pop_75, cluster = TRUE)
```

### Hausdorff

```{r}
d_haus <- region_distance_matrix(
  regions         = batches_proj,
  label_col       = "BatchName",
  method          = "hausdorff",
  units           = "km"
)
plot_distance_heatmap(d_haus, cluster = TRUE)
```

### Extended Hausdorff

```{r}
#| eval: false
## not yet implemented. need method = exthausdorff to be implemented in region_distance_matrix
d_haus_50 <- region_distance_matrix(
  regions         = batches_proj,
  label_col       = "BatchName",
  method          = "exthausdorff", ## todo: implement
  #tracts          = tracts_proj, ## don't need for ext haus actually
  units           = "km",
  perc_area       = 0.5 ## todo: implement 
)
plot_distance_heatmap(d_haus_50, cluster = TRUE)
```

## Correlation between distances

```{r}
dists_wide <- data.frame(
  dists_centroid = as.numeric(d_centroid[upper.tri(d_centroid)]),
  d_pop_25       = d_pop_25[upper.tri(d_pop_25)],
  d_pop_50       = d_pop_50[upper.tri(d_pop_50)],
  d_pop_75       = d_pop_75[upper.tri(d_pop_75)],
  d_haus = as.numeric(d_haus[upper.tri(d_haus)])
)

panel_with_abline <- function(data, mapping, ...) {
  ggplot(data, mapping) +
    geom_point(alpha = 0.6) +
    geom_smooth(method = "lm", se = FALSE) +
    geom_abline(slope = 1, intercept = 0, color = "red", linetype = "22")
}

GGally::ggpairs(
  dists_wide,
  lower = list(continuous = panel_with_abline)
)
```
