---
title: "RSV Bayes Renewal — Agreement Plots (Variant A vs Variant B)"
author: "Jose Palacio"
format: html
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

# What this file does

Reproduces the two-panel scatter figure comparing the posterior means of
$R_t$ and $I_t$ between the two observation-noise specifications of the
Ensor/Palacio single-plant renewal model:

- **Variant A** (`sigma_mode = "1byplant"`): homoscedastic viral-load
  observation noise with a single scalar $\hat\sigma_y$ per plant.
- **Variant B** (`sigma_mode = "filter"`): heteroscedastic input using
  the week-specific filter uncertainty $\hat P_t$ from the MARSS
  Stage-1 filter.

Each dot is one week of a given plant; the dashed line is the identity
$y = x$. Points close to the line mean Variant A and Variant B agree
on that week.

Assumes you have already run the main renewal QMD (`Bayes_Renewal_Code_Ensor.qmd`)
and saved the two fits as `.rds` files (see the "Save fits to disk" chunk
of that file):

```
outputs/fitA_cpd_1byplant.rds
outputs/fitB_cpd_filter.rds
```

If you have not saved them yet, just source the fitting chunks of the
main QMD first, then run this file.

```{r}
#| label: packages
suppressPackageStartupMessages({
  library(tidyverse)
  library(patchwork)
})
```

# 1 · Paths

```{r}
#| label: paths
proj_root <- if (Sys.info()["sysname"] == "Darwin") {
  "~/Documents/Claude/Projects/NOTS/rsv_mcmc_stan_geoW"
} else {
  normalizePath(".", mustWork = FALSE)
}
proj_root <- normalizePath(proj_root, mustWork = TRUE)

FIT_A_PATH <- file.path(proj_root, "outputs", "fitA_cpd_1byplant.rds")
FIT_B_PATH <- file.path(proj_root, "outputs", "fitB_cpd_filter.rds")

stopifnot(file.exists(FIT_A_PATH), file.exists(FIT_B_PATH))
OUT_DIR <- file.path(proj_root, "outputs", "figures")
dir.create(OUT_DIR, showWarnings = FALSE, recursive = TRUE)
```

# 2 · Load the two fits

```{r}
#| label: load-fits
fitA <- readRDS(FIT_A_PATH)   # Variant A: sigma_y homoscedastic per plant
fitB <- readRDS(FIT_B_PATH)   # Variant B: filter (P_t) as input

# Sanity: same P and T?
stopifnot(dim(fitA$y_PT) == dim(fitB$y_PT))
P  <- nrow(fitA$y_PT)
Tt <- ncol(fitA$y_PT)
cat(sprintf("Loaded fits: P = %d plants, T = %d weeks\n", P, Tt))
```

# 3 · Extract posterior means of R_t and I_t per (plant, week)

Each fit exposes `extract_array(name)` from the `renewal_model()` return
object (see main QMD, section 10). The array has dims `[draws, P, T]`;
we collapse over draws to get posterior means.

```{r}
#| label: extract-posterior-means

posterior_mean_PT <- function(fit, param) {
  arr <- fit$extract_array(param)      # [draws, P, T]
  if (is.null(arr))
    stop("Parameter '", param, "' not found in fit.")
  apply(arr, c(2, 3), mean, na.rm = TRUE)
}

Rt_A <- posterior_mean_PT(fitA, "Rt")   # P x T
Rt_B <- posterior_mean_PT(fitB, "Rt")

It_A <- posterior_mean_PT(fitA, "It")
It_B <- posterior_mean_PT(fitB, "It")

# Sanity
stopifnot(dim(Rt_A) == c(P, Tt), dim(Rt_B) == c(P, Tt))
cat(sprintf("Rt_A range: [%.3f, %.3f]\n", min(Rt_A), max(Rt_A)))
cat(sprintf("Rt_B range: [%.3f, %.3f]\n", min(Rt_B), max(Rt_B)))
cat(sprintf("It_A range: [%.1f, %.1f]\n", min(It_A), max(It_A)))
cat(sprintf("It_B range: [%.1f, %.1f]\n", min(It_B), max(It_B)))
```

# 4 · Build long-form data frames for ggplot

```{r}
#| label: long-form-data
# Plant labels: if you have specific plant codes/names, replace here.
# For the single-plant Ensor 69th Street case, P = 1 and plant_label = "69".
plant_labels <- if (P == 1) "69" else paste0("Plant", seq_len(P))

# Reshape wide P x T matrices into long tibbles with columns
# (plant, week, valA, valB) — one row per (plant, week).
mats_to_long <- function(mat_A, mat_B, value_name = "value") {
  tibble(
    plant = rep(plant_labels, times = Tt),
    week  = rep(seq_len(Tt), each = P),
    A     = as.vector(mat_A),
    B     = as.vector(mat_B)
  ) |>
    rename(!!paste0(value_name, "_A") := A,
           !!paste0(value_name, "_B") := B)
}

df_Rt <- mats_to_long(Rt_A, Rt_B, "R")
df_It <- mats_to_long(It_A, It_B, "I")
```

# 5 · The two scatter plots

```{r}
#| label: scatter-plots
#| fig-width: 12
#| fig-height: 5.5

# Common theme matching the paper style
plot_theme <- theme_minimal(base_size = 11) +
  theme(
    plot.title      = element_text(hjust = 0.5, face = "plain"),
    strip.background = element_rect(fill = "grey95", color = "grey60"),
    strip.text      = element_text(size = 10),
    axis.title      = element_text(size = 10),
    panel.grid.minor = element_blank()
  )

# Panel A: R_t agreement (linear axes)
p_Rt <- ggplot(df_Rt, aes(x = R_B, y = R_A)) +
  facet_wrap(~ plant, scales = "free", nrow = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black") +
  geom_point(alpha = 0.55, color = "grey30", size = 1.3) +
  labs(title = expression(R[t]),
       x = expression("Filter Estimates, "*hat(P)[t]),
       y = expression("Viral Load Measurements, "*sigma[y])) +
  plot_theme

# Panel B: I_t agreement (log-scale, since I spans orders of magnitude)
p_It <- ggplot(df_It, aes(x = I_B, y = I_A)) +
  facet_wrap(~ plant, scales = "free", nrow = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black") +
  geom_point(alpha = 0.55, color = "grey30", size = 1.3) +
  scale_x_log10() + scale_y_log10() +
  labs(title = expression(I[t]),
       x = expression("Filter Estimates, "*hat(P)[t]),
       y = expression("Viral Load Measurements, "*sigma[y])) +
  plot_theme

# Combine with letter labels A/B (matches paper style)
fig <- (p_Rt | p_It) +
  plot_annotation(tag_levels = "A") &
  theme(plot.tag = element_text(face = "plain", size = 12))

fig
```

# 6 · Save to PDF

```{r}
#| label: save-pdf
out_pdf <- file.path(OUT_DIR, "CPL_vs_CPD_agreement.pdf")
ggsave(out_pdf, fig, width = 12, height = 5.5)
cat(sprintf("Saved: %s\n", out_pdf))
```

# 7 · Numerical agreement diagnostics (optional)

Beyond the visual, quantify agreement with correlations and mean
absolute error. Points close to the identity line correspond to high
correlations and low MAE.

```{r}
#| label: agreement-numerics
agreement_summary <- function(df, valA, valB, param_name) {
  df |>
    group_by(plant) |>
    summarise(
      n        = sum(is.finite(.data[[valA]]) & is.finite(.data[[valB]])),
      pearson  = cor(.data[[valA]], .data[[valB]], use = "pairwise.complete.obs"),
      MAE      = mean(abs(.data[[valA]] - .data[[valB]]), na.rm = TRUE),
      MAE_rel  = MAE / mean(.data[[valB]], na.rm = TRUE),
      .groups = "drop"
    ) |>
    mutate(quantity = param_name, .before = 1)
}

agreement <- bind_rows(
  agreement_summary(df_Rt, "R_A", "R_B", "R_t"),
  agreement_summary(df_It, "I_A", "I_B", "I_t")
)

print(agreement)
```

# Notes

- **Panel A vs Panel B**: If Variant A and Variant B agree on point
  estimates, all points lie on the dashed identity line. Off-diagonal
  scatter reflects Variant-A's homoscedastic scalar-noise assumption
  compensating with slightly different latent trajectories vs the
  time-varying-noise Variant B.
- **Log-scale in Panel B ($I_t$)**: $I_t$ spans several orders of
  magnitude within a season (~10 at baseline to >100 at peak); the
  log-scale makes the dispersion at small values visible without hiding
  the peak values.
- **Reproducibility**: if you re-run the fits with different MCMC seeds,
  point estimates shift by Monte Carlo error (~1-2%), but the visual
  agreement structure is stable.
